---
title: "Exploratory Analysis"
author: "Joel Parker"
date: "2023-07-17"
output: html_document
---
```{r setup, include=FALSE}
# Set global chunk obptions
knitr::opts_chunk$set(fig.width=12, fig.height=12)

# Data 
library(tidyverse)
library(openxlsx)
library(kableExtra)
library(table1)
library(flextable)
library(reshape2)


# Heat Map
library(pheatmap)
library(RColorBrewer)
library(rstatix)
library(plotly)

# PCA
library(factoextra)
library(FactoMineR)


## Combining plots
library(ggpubr)


## Image package
library(magick)

## Pairwise comparisons
library(emmeans)

# Metabolomics Pipeline source files
files.sources = list.files("../R")
sapply(paste0("../R/",files.sources), source)

```


## Exploratory Analysis

In data exploration, we use several methods to help us better understand the 
underlying patterns in the data without using a formal hypothesis test. In this
pipeline, we are going to focus on two methods for data exploration:

A.) Principal component analysis

B.) Heat maps

Before diving into the data exploration, we need read the analysis data set. Note,
we use check.names=F since column names of the analysis data are numeric. 

```{r load-data2}
# 1. Read analysis dataset
analysis_data <- read.csv("../Data/Processed/analysis_data.csv", check.names = F)
```

### Table 1
The goal of this section is to design a table that resembles the study design of
your experiment. This step allows us to see descriptive statistics of the samples
in our data. For this, we are using the analysis data to define a table with a similar
structure as the table below.

|         | Strat 1      |Strat 2    |
|:-------:|:------------:|:---------:|
| *Var1 * |XX     | XX   |  
| *Var 2* | XX    | XX   |
| *Var 3* | XX    | XX   |


Alternatively, we can define a table non-stratified table with only the number of
observations within each experimental group. The following code will create the
desired table 1 for both scenarios.


In this chunk, we will label the analysis data variable names. Labeling the variable
names will allow the tables to display your chosen variable names while maintaining
the original variable name within the data. We do this in two steps.

1. Choose the variable name.
  + Change the name assigned to var1 to match the name of the variable in the
  analysis data set which you want to include in the table.
  
  + Repeat this for as many variables as needed.
  
2. Choose a single variable to stratify the table by.
  
3. Assign the variable a label name
  + you can change the label of the variable in the table by relabeling the
  variable to the desired name on the right side of the equation. 
  
4. Assign a stratified variable a label name for the table. 

```{r table one labels}
# 1. Choose the meta data variable name.
var1 = "GROUP_NAME"

var2 = "TIME1"


# 2. Choose a single variable to stratify the table by.
stratified_var = "Gender"


# 3. Assign the variable a label name
label(analysis_data[,var1]) <- "Treatment Group" 

label(analysis_data[,var2]) <- "Time"


# 4. Assign a stratified variable a label name for the table. 
label(analysis_data[,stratified_var]) = "Gender"
```


Now that you have properly assigned the variables, we will create table 1. To
create the table, we will be using the table1 function. You must change the
arguments depending on what you want the table to display. For example:
* *Table 1 without interaction*: 

$$
table1(\sim \underbrace{var_1 + var_2 ...}_{\text{Variable names for rows}},   \ \text{data = analysis_data}  )
$$



* *Table 1 with interaction*

$$
table1(\sim \underbrace{var_1 + var_2 ...}_{\text{Variable names for rows}} \underbrace{| \ stratified\_var}_{\text{Statified Variable}},  \ \text{data = analysis_data} )
$$
Above is how you change the table1 function to display the table with and without
a stratified variable. In the next chunk, we utilize the following steps.

1. Create table 1. In this step you will need to use the same variable names as
in the above chunk within the table1 function. 

2. Display the table

3. Save the table in the folder "/Outputs/Tables/" with the file name "table1". 

```{r table one}
# 1. Creates table1
tbl1 <- table1(~ TIME1 + GROUP_NAME| Gender
               , data = analysis_data)


# 2. Displays table 1
tbl1


# 3. saves table 1 
t1flex(tbl1) %>%
  save_as_docx(path = paste0("../Outputs/Tables/","table1.docx"))
```


###  Principal Component Analysis

In general, Principal component analysis (PCA) reduces the number of variables 
in a dataset while preserving as much information from the data as possible. At 
a high level, PCA is constructed such that the first principal component accounts
for the largest possible amount of variance within the data. The second principal
component accounts for the largest amount of remaining variance, and so on. 
Additionally, each of the principal components produced by PCA is uncorrelated 
with each of the other principal components. At a high level, PCA allows us to 
visualize sources of variation in the data.

Here, we will graph the first two principal components of our dataset. In the 
PCA plot, we can label each point based on variables from the metadata.

The following chunk runs PCA on the scale_peak_data matrix in the following 
steps:

1.Create a vector of non metabolite variables that are in the the analysis data
set. These variables include information about the experimental conditions. 

2. Create a data set for the principle component analysis which only includes 
information on the metabolites. 

3. Run PCA of the pca_dat matrix containing only the metabolites. 

4. Define the plot labels. This is a character string matching the name of one of 
the variables in the analysis_data which will be used to label points on the PCA
graph. 

5. Create and display the PCA plot.

6. Save the PCA plot in the "Outputs/Plots" folder with the file name PCA.pdf


```{r pca plots}
# 1. Create the non-metabolite vector. 
non_metabolite <- c("PARENT_SAMPLE_NAME",
                             "GROUP_NAME",
                             "TIME1",
                             "Gender")


# 2. Create PCA data containing only metabolite data
pca_dat <- analysis_data[,!names(analysis_data) %in% non_metabolite]


# 3. Run PCA of the pca_dat matrix containing only the metabolites.   
res.pca <- PCA(pca_dat, 
               graph = F)


# 4. Define labels
meta_var = "Gender"


# 5. Create figure 
fviz_pca_ind(res.pca, 
             label = "none",
             habillage = as.factor(analysis_data[,meta_var])) 


# 6. Save figure
ggsave(paste0("../Outputs/Plots/","PCA.pdf"), width = 10, height = 10)
```


Suppose you notice a variable with clearly separated groups, and is not a 
variable of interest. In that case, you may want to consider stratifying your 
analysis downstream by the values of that variable. For example, if we are 
looking at the effects of a specific drug, and we notice in the above plot that 
the samples are grouped by sex, then we may want to consider stratifying the 
analysis by male and female.

### Heatmaps

Another exploratory analysis tool we can use is heatmaps, whih is a gridded plot
based on the x-axis- and y-axis labels. The color of the spot on the grid is based on the 
value's intensity. For our heatmap, the x-axis will be the samples, and the y-axis
will be the metabolites. The values determining the colors will be the log 
peak value for each chemical in each observation. We can order our observations 
to see intensity patterns based on our experimental conditions. This is another
way of visualizing sources of variation within our data. To do this, we 
need to merge the chemical annotation, meta, and scaled peak data. Then we need 
to arrange the data based on our experimental conditions. The create_heatmap_Data
function in the R folder will help us prepare the data for the heatmap. To do this
the create_heatmap_Data function takes three arguments.


* Analysis data (analysis_data) 

* Heatmap Variables (heatmap_variables) - defines the variables used to order
the samples.

* ... -  


The main utility of create_heatmap data is in (…), which allows you to arrange 
the data based on experimental conditions. If you are familiar with dplyr, the 
arrange function orders samples based on the arguments passed into (...). If you 
are unfamiliar with dplyr, an overview of the arrange function is [here](https://dplyr.tidyverse.org/reference/arrange.html).


We will first look at the heatmap with all metabolites. The resulting heatmap
saves into the Outputs/Plots folder under the filename "AllMetabolites.pdf."

In this chunk we complete the following steps. 

1. Define the heatmap_variables that will be used to order the heatmap

2. Run the create_heatmap_Data function defined in the R script folder. This function
is going to produce the matrices needed for the heatmap

3. Create a palette for the heatmap. By default we are using a red/blue palette.

4. Create the vals and labels matrices that will define the values of the heatmap
and the labels of the heatmap. 

5. Display heatmap

6. Save heatmap in the "Outputs/Plots" folder under the filename "Heatmap_all"



```{r overall heatmap}
# 1. define meta analysis variables
heatmap_variables <-   c("GROUP_NAME",
                             "TIME1",
                             "Gender")


# 2. Create heatmap data
heatmap_data <- create_heatmap_Data(analysis_data,heatmap_variables = heatmap_variables ,
                               Gender, GROUP_NAME, desc(TIME1)) # Arranges data frame for heatmap (...)



# 3. Heat map colors 
palette <- colorRampPalette(rev(brewer.pal(10, "RdBu")))(256)



# 4. Values for heatmap
vals = heatmap_data$heatmap_data_vals

meta = heatmap_data$heatmap_variables


# 5. Create heatmap and save. 
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA)


# 6. Save heatmap
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA,filename = paste0("../Outputs/Plots/","Heatmap_all.pdf"))

```



#### Heatmap with top 50 metabolites 

Looking at the heatmap created in the previous chunk with all of the metabolites
can make it challenging to determine which line goes with which metabolite. To 
overcome this, we can filter the metabolites based on which metabolites have the
highest mean log-scaled peak data across all observations. In the heatmap below,
we select the top 50 metabolites. Once we know the top 50 metabolites with the
highest mean value, we can filter the peak_data_log data only to include those 
metabolites. Then we can run the create_heatmap function again and create the 
heatmap data. To do this the following steps are completed:

1. Define the heatmap_variables to use in the heatmap

2. Select the top 50 metabolites from the analysis data based off of mean value.

3. Filter the analysis_data to only include the top 50 metabolites

4. Generate heatmap data using the create_heatmap_Data function

5. Create a palette for the heatmap. By default, we are using a red/blue palette.

6. Create the vals and the labels that will define the values of the heatmap
and the labels of the heatmap. 

7. Display heatmap

8. Save heatmap in the "Outputs/Plots" folder under the filename "HeatmapTop50Metabolites"


```{r top heatmap}
# 1. Define meta analysis variables
heatmap_variables <- c("GROUP_NAME", "TIME1", "Gender")


# 2. Find the top 50 metabolites
select_variables <- analysis_data %>% 
  select(-all_of(c("PARENT_SAMPLE_NAME",heatmap_variables))) %>%
  summarise(across(everything(),\(x) mean(x,na.rm = T))) %>%
  pivot_longer(cols = everything()) %>%
  arrange(desc(value)) %>%
  slice(c(1:50))


# 3. Filter to the top 50 metabolites
analysis_data_top50 <- analysis_data %>%
  select(all_of(c("PARENT_SAMPLE_NAME", heatmap_variables)), select_variables$name) 
  

# 4. Create heatmap data
heatmap_data <- create_heatmap_Data(analysis_data_top50,
                                    heatmap_variables = heatmap_variables ,
                               Gender, GROUP_NAME, desc(TIME1)) # Arranges data frame for heatmap (...)


# 5. Heat map colors 
palette <- colorRampPalette(rev(brewer.pal(10, "RdBu")))(256)


# 6. Values for heatmap
vals = heatmap_data$heatmap_data_vals

meta = heatmap_data$heatmap_variables


# 7. Create heatmap and save. 
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA)


# 8. Save heatmap
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA,filename = paste0("../Outputs/Plots/","HeatmapTop50Metabolites.pdf"))




```



