---
title: "Metabolon Work Flow"
output:
  html_document: 
    toc: true
    df_print: paged
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
      out_dir <- "../Outputs/Reports";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---



```{r setup, include=FALSE}
# Set global chunk obptions
knitr::opts_chunk$set(fig.width=12, fig.height=12)

# Data 
library(tidyverse)
library(openxlsx)
library(kableExtra)
library(table1)
library(data.table)
library(flextable)


# Heat Map
library(pheatmap)
library(RColorBrewer)
library(rstatix)



# PCA
library(factoextra)
library(FactoMineR)


## Combining plots
library(ggpubr)

## Image package
library(magick)
# Source R scripts 
source("../R/Metabolomics_R_code.R")
```

## Introduction
The purpose of this pipeline is to provide additional analysis to complement the
analysis by Metabolon, where they identified biochemicals (metabolites) that 
differed significantly across experimental groups. In the Metabolon data, each 
"sub-pathway" classifies a collection of metabolites, and each "super-pathway" 
classifies a collection of sub-pathways. Hypothesis testing at the sub-pathway 
level can provide valuable insight into how the experimental conditions affect 
metabolic profiles.
This pipeline walks through how to perform hypothesis testing on the metabolomics
data at the sub-pathway level and provides useful tables and figures to aid our 
understanding. The below figure highlights the steps we will be taking, which 
highlight the main sections of this pipeline by: 

1. Normalization and Standardization

2. Data Exploration

3. Subpathway Analysis

4. Pairwise comparisons

5. Plots

```{r workflow}
img = image_read("../Workflow.png")

image_trim(img)
```

The R "chunks" separate each of the analysis steps within the main sections of 
the pipeline. In some chunks, there may need to be lines of code that need to 
change from experiment to experiment. This pipeline provides a list of steps for
each chunk and highlights places where you need to make changes. 

## Link and folder set up

To get started, in the Metabolomics pipeline working directory, there are five 
folders with the following folder structure: 

-   Data
    - Metabolon: Contains the .xlsx file from Metabolon and other files related 
    to the metadata.
    - Processed: Contains data processed from this pipeline. 
-   Code: 
-   Outputs
    -   Plots
    -   Tables
    -   Reports
- R: Contains and .R file with additional functions used for this pipeline.
- renv: This folder is generated by renv to restore package versions.

You should save the .xlsx file from Metabolon and any additional metadata files 
to the Data/Metabolon folder. The main output folder will contain the plots and 
tables generated from this pipeline. Additionally, you can run this pipeline as 
a report and save it as a .pdf or .html. The report saves to the "Outputs/Reports"
folder.



## Data loading

We will use the Sample Meta Data, Peak Area Data, and the chemical annotation 
provided in the Data Tables Excel file from Metabolon for the analysis. We can 
also add Sample Meta Data not captured in the Metabolon Excel file. Therefore, 
we will need links to the following files

1.  Metabolon Excel file
2.  Additional sample meta data (with a unique ID for each sample
    allowing us to link it to each observation)
    
If there are no additional variables, then you can set the additional_vars
value to NULL.

```{r data links}
# 1. Metabolome excel file 
met_excel <- "../Data/Metabolon/UNAZ-0501-22VW_ DATA TABLES.xlsx"


# 2. Path to additional variables. Please make this NULL if there are not any additional
# Files
additional_vars = "../Data/Metabolon/AdditionalVars.xlsx"
```

Once you provide a path to the data, we can load the data into R. This chunk 
will access the Metabolon Excel file specified above and pull out three data sets
within the Excel file. Below is a table that shows the data that we pull from the
Excel file and the tab within the Metabolon Excel sheet.

| Data Accessed       | Tab                 |
|:-------------------:|:-------------------:|
| Raw Peak            | Peak Area Data      |
| Meta Data           | Sample Meta Data    |
| Chemical Annotation | Chemical Annotation |

If the names of the tabs containing this data are different in your Excel file, 
there are two solutions. You can change the name of the tabs to match those 
listed above, or you can replace the "sheet" variable in the "read.xlsx" 
function to match the naming convention within your Metabomon Excell workbook.

In the following chunk we are:

1. Loading the peak data from the Metabolon .xlsx file. 

2. Load the meta data from the Metabolon .xlsx file. If any additional variables
are provided in the links above, these variables will be added to the meta data. 

3. Loading in the chemical annotation data. 
```{r load data}
# 1. load peak data Peak area data
peak_data <- read.xlsx(met_excel,
                sheet = "Peak Area Data")


# 2. Load metadata with no additional variables
if(is.null(additional_vars)){
    meta_data <-  read.xlsx(met_excel,
                             sheet = "Sample Meta Data")
}   


# Load meta_data with additional variables
if(!is.null(additional_vars)){
    meta_data <-  read.xlsx(met_excel,
                             sheet = "Sample Meta Data")

    additional <- read.xlsx(additional_vars)
    
    meta_data <- merge(additional
                       ,meta_data, by="PARENT_SAMPLE_NAME")
}


# 3. Chemical names key
chem_data <- read.xlsx(met_excel,
                       sheet = "Chemical Annotation")
```


After reading the data, we want to inspect the loaded data to ensure everything 
is stored properly. In the following chunk we will view the first 6 rows of the 
metadata, chemical annotation data, and the first 6 row and ten columns of the 
peak data. Therefore, the steps we are taking are,

1. First 6 rows of metadata

2. First 6 rows of the chemical annotation data

3. First 6 rows and 10 columns of the peak data. 

```{r check data frames}
# 1. First 6 rows of meta data
head(meta_data) 


# 2. First 6 rows of the chemical annotation data
head(chem_data)


# First 6 rows and 10 columns of the peak data. 
head(peak_data[,1:10])
```

### Table 1

The goal of this section is to design a table that resembles the study design of
your experiment. This step allows us to see descriptive statistics of the samples
in our data. For this, we are using the metadata to define a table with a similar
structure as the table below.

|         | Strat 1      |Strat 2    |
|:-------:|:------------:|:---------:|
| *Var1 * |XX     | XX   |  
| *Var 2* | XX    | XX   |
| *Var 3* | XX    | XX   |


Alternatively, we can define a table with only the number of observations within
each experimental group. The following code will create the desired table 1 for
both scenarios.


In this chunk, we will label the metadata's variable names. Labeling the variable
names will allow the tables to display your chosen variable names while maintaining
the original variable name within the data. We do this in two steps.

1. Choose the metadata variable name.
  + Change the name assigned to var1 to match the name of the variable in the
  data set which you want to include in the table
  
  + Repeat this for as many variables as needed.
  
2. Choose a single variable to stratify the table by.
  
3. Assign the variable a label name
  + you can change the label of the variable in the table by relabeling the
  variable to the desired name on the right side of the equation. 
  
4. Assign a stratified variable a label name for the table. 

```{r table one labels}
# 1. Choose the meta data variable name.
var1 = "GROUP_NAME"

var2 = "TIME1"


# 2. Choose a single variable to stratify the table by.
stratified_var = "Gender"


# 3. Assign the variable a label name
label(meta_data[,var1]) <- "Treatment Group" 

label(meta_data[,var2]) <- "Time"


# 4. Assign a stratified varibale a label name for the table. 
label(meta_data[,stratified_var]) = "Gender"
```


Now that you have properly assigned the variables, we will create table 1. To
create the table, we will be using the table1 function. You must change the
arguments depending on what you want the table to display. For example:
* *Table 1 without interaction*: 

$$
table1(\sim \underbrace{var_1 + var_2 ...}_{\text{Variable names for rows}},   \ \text{data = meta_data}  )
$$



* *Table 1 with interaction*

$$
table1(\sim \underbrace{var_1 + var_2 ...}_{\text{Variable names for rows}} \underbrace{| \ stratified\_var}_{\text{Statified Variable}},  \ \text{data = meta_data} )
$$
Above is how you change the table1 function to display the table with and without
a stratified variable. In the next chunk, we utilize the following steps.

1. Create table 1. In this step you will need to use the same variable names as
in the above chunk within the table1 function. 

2. Display the table

3. Save the table in the folder "/Outputs/Tables/" with the file name "table1". 

```{r table one}
# 1. Creates table1
tbl1 <- table1(~ TIME1 + GROUP_NAME| Gender
               , data = meta_data)


# 2. Displays table 1
tbl1


# 3. saves table 1 
t1flex(tbl1) %>%
  save_as_docx(path = paste0("../Outputs/Tables/","table1.docx"))
```


## Normalization and Standardization
In its raw form, the peak data contain counts for each metabolite in each sample.
Standardization and normalization are important steps to improve the 
signal-to-noise ratio. In the next section, we are implementing the same 
standardization and normalization methods used by Metabolon.

### Missing Peak data
As part of the Metabolon workflow, the raw peak data goes through a series of 
processing steps to prepare the data for analysis. The first part of this process
involves handling the missing peak data in two stages.

1.) Remove chemicals with more than 50% of the observations missing.

2.) For the remaining chemicals with missing data, impute the missing values with
1/5 of the minimum positive value. 


For the first step, we will summarize the missingness for the metabolites in a 
table to show us which metabolites will exclude in the analysis. In the following
chunk we:


1. Create a data set with  the number of missing observations for each metabolite. 
As part of this step, we are also filtering so that only metabolites are included
that have more than 50% of the observations missing. 

2. Filter the columns of the peak data to exclude the metabolites identified in step
1. 


3. Create a table that shows the metabolites excluded with the proportion of missing
observations for each metabolite. 

4. Display the table of excluded metabolites

5. Save table in the "Outputs/Tables" folder with the filename "MissingChemicals"


```{r remove missing features}
# 1. Create a dataset with  the number of missing observations for each metabolite. 
missing_variables <- peak_data %>%
  mutate(across(everything(), ~ifelse(is.na(.), 1, 0))) %>%
  summarise_all(sum) %>%
  pivot_longer(cols = everything(), names_to = "CHEM_ID", values_to = "number_missing") %>%
  left_join(select(chem_data, CHEM_ID, CHEMICAL_NAME) %>%
              mutate(CHEM_ID = as.character(CHEM_ID)),
            by = "CHEM_ID") %>%
  arrange(number_missing) %>%
  filter(number_missing > 0.5*(nrow(peak_data)))


# 2. Filter the columns of the peak data to exclude the metabolites identified in step 1.
peak_data_filter <- peak_data %>%
  select(-missing_variables$CHEM_ID)


# 3. Create a table that shows the metabolites excluded
miss_tabl<- missing_variables %>%
              select(CHEMICAL_NAME, number_missing) %>%
              mutate( prop_missing = round(number_missing/nrow(peak_data), digits = 3)) %>%
              flextable() %>%
              set_header_labels(CHEMICAL_NAME="Chemical Name", number_missing="Number Missing", number_missing = "Proportion Missing") %>%
              set_table_properties(layout = "autofit") %>%
              theme_vanilla()


# 4. Display Table
miss_tabl


# 5. Save table
save_as_docx(miss_tabl,path = "../Outputs/Tables/MissingChemicals.docx")

```


Once we remove the metabolites with more than 50% of the observations missing, 
we impute 1/5 of the minimum value for the remainder with missing values. We do 
this in the following steps.

1. Initialize the new peak_data_imputed matrix

2. Find the minimum value for each metabolite and compute 1/5 of that value

3. Create a for loop such that if a metabolite has a missing observation, then
1/5 of the minimum value for that metabolite is imputed. 


```{r replace missing observations}
# 1. Initialize the new peak_data_imputed matrix
peak_data_imputed <- peak_data_filter


# 2. Find the minimum value for each metabolite and compute 1/5 of that value
peak_data_mins <- peak_data_imputed %>%
  select(-PARENT_SAMPLE_NAME) %>%
  summarise_all(min, na.rm = T) %>%
  mutate(across(everything(), ~(./5)))


# 3. Impute the value
for(i in colnames(peak_data_mins)){
  if(length(peak_data_imputed[,i][is.na(peak_data_imputed[,i])]) > 0){
    peak_data_imputed[which(is.na(peak_data_imputed[,i])),i] <- as.numeric(peak_data_mins[i])
  }
}
```


### Normalization

Once we impute the missing data, the Metabolon workflow follows a three-step 
process for normalization.

A. Each metabolite's peak data is standardized by the median value of the 
metabolites observations. 

B. Each observation is log10 transformed

C. Each chemical is autoscaled by subtracting the mean from each value and
dividing by the standard deviation. 

The next two chunks will perform these three steps to produce a dataset called 
"peak_data_scale". Peak_data_scale will be the data set used for data exploration
in the next section.

In the first chunk we:

1. Initialize a new peak_data_norm matrix

2. Create a matrix containing the median value for each metabolite 

3. Divide each value for each metabolite by the median value of that metabolite 

4. Log10 transform all of the values


```{r median normalization}
# 1. Initialize a new peak_data_norm matrix
peak_data_norm <- peak_data_imputed 


# 2. Create a matrix containing the median value for each metabolite 
peak_data_med <- peak_data_norm %>%
  select(-PARENT_SAMPLE_NAME) %>%
  summarise_all(median, na.rm = T)


# 3. Divide each value for each metabolite by the median value of that metabolite
for(i in colnames(peak_data_med)){
  peak_data_norm[,i] <- peak_data_norm[,i]/peak_data_med[,i]
}


# 4. Log10 transform all of the values
peak_data_log <- peak_data_norm %>%
  mutate_if(is.numeric, log10)
```


For the next chunk, we "auto-scale" the features. We will do this in two steps:

1. Initialize the new matrix peak_data_scale, this is the matrix that we will use
in down stream tasks. 

2. Auto-scale by subtracting the mean and dividing by the standard deviation of
each metabolite. 


```{r autoscale variables}
# 1. Initialize the new matrix peak_data_scale
peak_data_scale <- peak_data_log


# 2. Auto-scale
for(i in colnames(peak_data_log)[-1]){
  peak_data_scale[,i] <- (peak_data_scale[,i]-mean(peak_data_scale[,i]))/sd(peak_data_scale[,i])
}
```


### Save data
Now that we have imputed and normalized our data, we need to save the data for subsequent sections. The data which we will use downstream are:

1. Meta data (meta_data)

2. Peak Scaled Data (peak_data_scale)

3. Chemical Annotation data (chem_data)

We will save these data sets to the "Data/Processed" folder. 

```{r save-data 1}

# 1. Save meta data
write.csv(meta_data, file = "../Data/Processed/meta_data.csv", row.names = F)


# 2. Save peak scaled data
write.csv(peak_data_scale, "../Data/Processed/peak_data_scale.csv",row.names = F)


# 3. save chemical annotaion data
write.csv(chem_data, file = "../Data/Processed/chem_data.csv", row.names = F)

```


## Data Exploration

In data exploration, we use several methods to help us better understand the 
underlying patterns in the data without using a formal hypothesis test. In this
pipeline, we are going to focus on two methods for data exploration:

A.) Principal component analysis

B.) Heatmaps

Before diving into the data exploration, we need to read the processed data from
the Normalization and Standardization Section. We will need to read the following
files

1. Meta data: We will need to call this meta_data by default for the other down-
stream tasks. If you make any changes to this variable name then you will have to
make changes in down-stream chunks.

2. Scales peak data: We will need to call this peak_data_scale by default for the
other down stream-tasks. If you make any changes to this variable name then you 
will have to make changes in down-stream chunks.

3. Chemical annotation data: We will need to call this chem_data by default for 
the other down-stream tasks. If you make any changes to this variable name then 
you will have to make changes in down-stream chunks.


```{r load-data2}
# 1. Read meta-data
meta_data <- read.csv("../Data/Processed/meta_data.csv", check.names = F)

# 2. Read scaled peak data
peak_data_scale <- read.csv("../Data/Processed/peak_data_scale.csv", check.names = F)

# 3. Read chemical annotations
chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)
```


###  Principal Component Analysis

In general, Principal component analysis (PCA) reduces the number of variables 
in a dataset while preserving as much information from the data as possible. At 
a high level, PCA is constructed such that the first principal component accounts
for the largest possible amount of variance within the data. The second principal
component accounts for the largest amount of remaining variance, and so on. 
Additionally, each of the principal components produced by PCA is uncorrelated 
with each of the other principal components.

Here, we will graph the first two principal components of our dataset. In the 
PCA plot, we can label each point based on variables from the metadata.

The following chunk runs PCA on the scale_peak_data matrix in the following 
steps:

1. Set the meta analysis variables from the metadata that will be used 
to categorize points on the PCA plot. You must always include the variable
name that maps the observations from the scaled_peak_data to the meta_data. This
is typically the variable name "PARENT_SAMPLE_NAME". 

2. Merge the peak_data_scale and the meta_data matrices together to define the 
pca_dat

3. Create a list of all of the metabolite names in the newly generated pca_dat
matrix. 

4. Run PCA of the pca_dat matrix containing only the metabolites. 

5. Define the plot labels. This is a character string matching the name of one of 
the variables in the meta data. 

6. Create and display the PCA plot

7. Save the PCA plot in the "Outputs/Plots" folder with the file name PCA.pdf


```{r pca plots}
# 1. Set the meta analysis variables 
meta_analysis_variables <- c("PARENT_SAMPLE_NAME",
                             "GROUP_NAME",
                             "TIME1",
                             "Gender")


# 2. Merge the peak_data_scale and the meta_data matrices together
pca_dat <- peak_data_scale %>%
  left_join(meta_data[,meta_analysis_variables], 
            by = "PARENT_SAMPLE_NAME") %>%
  select(all_of(meta_analysis_variables),everything())


# 3. Create a list of all of the metabolite names
chems <- names(pca_dat)[(names(pca_dat)%in% chem_data$CHEM_ID)]


# 4. Run PCA of the pca_dat matrix containing only the metabolites.   
res.pca <- PCA(pca_dat[chems], 
               graph = F)


# 5. Define labels
meta_var = "Gender"


# 6. Create figure 
fviz_pca_ind(res.pca, 
             label = "none",
             habillage = as.factor(pca_dat[,meta_var])) 

# 7. Save figure
ggsave(paste0("../Outputs/Plots/","PCA.pdf"), width = 10, height = 10)
```


Suppose you notice a variable with clearly separated groups, and is not a 
variable of interest. In that case, you may want to consider stratifying your 
analysis downstream by the values of that variable. For example, if we are 
looking at the effects of a specific drug, and we notice in the above plot that 
the samples are grouped by sex, then we may want to consider stratifying the 
analysis by male and female.

### Heatmaps

Another exploratory analysis tool we can use is heatmaps, a gridded plot based on
the x-axis- and y-axis labels. The color of the spot on the grid is based on the 
value's intensity. For our heatmap, the x-axis will be the samples, and the y-axis
will be the metabolites. The values determining the colors will be the log10 
scaled peak value for each chemical in each observation. We can order our observations 
to see intensity patterns based on our experimental conditions. To do this, we 
need to merge the chemical annotation, meta, and scaled peak data. Then we need 
to arrange the data based on our experimental conditions. We have developed an R
function (create_heatmap) that does this in one step. The create_heatmap function
takes the following arguments.


* Peak scaled data (peak_data_scale)
* Chemical Annotation data (chem_data)
* Meta data (meta_data)
* Meta analyis variables
* ... 

The first three arguments of the function are data frames (data frame names are 
in parenthesis). The meta-analysis variables are specified in the
meta_analysis_variables, which contains all of the variables from the metadata 
that you would like to include in the heatmap.

The main utility of create_heatmap data is in (…), which allows you to arrange 
the data based on experimental conditions. If you are familiar with dplyr, the 
arrange function orders samples based on the arguments passed into (...). If you 
are unfamiliar with dplyr, an overview of the arrange function is [here](https://dplyr.tidyverse.org/reference/arrange.html).


We will first look at the heatmap with all metabolites. The resulting heatmap
saves into the Outputs/Plots folder under the filename "AllMetabolites.pdf."

In this chunk we complete the following steps. 

1. Define the meta analysis variables that will be used to order the heatmap

2. Run the create_heatmap function defined in the R script folder. This function
is going to produce the matrices needed for the heatmap

3. Create a palette for the heatmap. By default we are using a red/blue palette.

4. Create the vals and meta matrices that will define the values of the heatmap
and the labels of the heatmap. 

5. Display heatmap

6. Save heatmap in the "Outputs/Plots" folder under the filename "Heatmap_all"



```{r overall heatmap}
# 1. define meta analysis variables
meta_analysis_variables <- meta_analysis_variables <- c("PARENT_SAMPLE_NAME",
                             "GROUP_NAME",
                             "TIME1",
                             "Gender")


# 2. Create heatmap data
heatmap_data <- create_heatmap(peak_data_scale,chem_data,meta_data, meta_analysis_variables,
                               Gender, GROUP_NAME, desc(TIME1)) # Arranges data frame for heatmap (...)



# 3. Heat map colors 
palette <- colorRampPalette(rev(brewer.pal(10, "RdBu")))(256)



# 4. Values for heatmap
vals = heatmap_data$heatmap_data_vals

meta = heatmap_data$heatmap_meta_data


# 5. Create heatmap and save. 
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA)


# 6. Save heatmap
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA,filename = paste0("../Outputs/Plots/","Heatmap_all.pdf"))






```

#### Heatmap with top 50 metabolites 

Looking at the heatmap created in the previous chunk with all of the metabolites
can make it challenging to determine which line goes with which metabolite. To 
overcome this, we can filter the metabolites based on which metabolites have the
highest mean log-scaled peak data across all observations. In the heatmap below,
we select the top 50 metabolites. Once we know the top 50 metabolites with the
highest mean value, we can filter the peak_data_log data only to include those 
metabolites. Then we can run the create_heatmap function again and create the 
heatmap data. To do this the following steps are completed:

1. Define the meta analysis variables to use in the heatmap

2. Select the top 50 metabolites from the scaled peak data

3. Filter the scaled_peak_data to only include the top 50 metabolites

4. Generate heatmap data using the create_heatmap function

5. Create a palette for the heatmap. By default, we are using a red/blue palette.

6. Create the vals and meta matrices that will define the values of the heatmap
and the labels of the heatmap. 

7. Display heatmap

8. Save heatmap in the "Outputs/Plots" folder under the filename "HeatmapTop50Metabolites"


```{r top heatmap}
# 1. Define meta analysis variables
meta_analysis_variables <- c("PARENT_SAMPLE_NAME","GROUP_NAME", "TIME1", "Gender")


# 2. Find the top 50 metabolites
select_variables <- peak_data_scale %>% 
  select(-PARENT_SAMPLE_NAME) %>%
  summarise(across(everything(), mean, na.rm = T)) %>%
  pivot_longer(cols = everything()) %>%
  arrange(desc(value)) %>%
  slice(c(1:50))


# 3. Filter to the top 50 metabolites
peak_data_scale_top50 <- peak_data_scale %>%
  select(PARENT_SAMPLE_NAME, select_variables$name) 
  

# 4. Create heatmap data
heatmap_data <- create_heatmap(peak_data_scale_top50 ,chem_data,meta_data, meta_analysis_variables,
                               Gender, GROUP_NAME, desc(TIME1)) # Arranges data frame for heatmap



# 5. Heat map colors 
palette <- colorRampPalette(rev(brewer.pal(10, "RdBu")))(256)


# 6. Values for heatmap
vals = heatmap_data$heatmap_data_vals

meta = heatmap_data$heatmap_meta_data


# 7. Create heatmap and save. 
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA)


# 8. Save heatmap
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA,filename = paste0("../Outputs/Plots/","HeatmapTop50Metabolites.pdf"))




```


#### Stratified Heatmap

We may also want to stratify our heatmap based on a specific variable within our
data set. To do this, we have created the filter_heatmap function, which takes 
the output from the create_heatmap function and filters the heatmap data. The 
filter_heatmap function takes three arguments.


* heatmap_data - List of data frames from the create_heatmap function.

* drop_vars - Any metadata variables in the heatmap data that you would like 
to drop from the analysis. For instance, if we are looking at only male samples we may
want to drop the gender variable because it is no longer informative. 

* (...) arguments for the dplyr filter function. This will allow you to filter
the data based on the stratification you want. For example: Gender=="Male"

In the chunk below we are going to produce a heatmap for a single strata by:

1. Defining meta analysis variables

2. Selecting the top 50 metabolites

3. Filter the scaled_peak_data on the top 50 metabolites

4. Generate the heatmap data with the create_heatmap function

5. Filter the generated heatmap data on a specific strata using the filter_heatmap
function.

6. Create a palette for the heatmap. By default we are using a red/blue palette.

7. Create the vals and meta matrices that will define the values of the heatmap
and the labels of the heatmap. 

8. Display heatmap

9. Save heatmap in the "Outputs/Plots" folder under the filename "HeatmapTop50MetabolitesStrat1".
This filename should be changed to reflect the name of the strata instead of "Strat1".


```{r stratified heatmap}
# define meta analysis variables
meta_analysis_variables <- c("PARENT_SAMPLE_NAME","GROUP_NAME", "TIME1", "Gender")
  


# Find the top 50 metabolites by log peak area
select_variables <- peak_data_scale %>% 
  select(-PARENT_SAMPLE_NAME) %>%
  summarise(across(everything(), mean, na.rm = T)) %>%
  pivot_longer(cols = everything()) %>%
  arrange(desc(value)) %>%
  slice(c(1:50))

# Filter to the top 50 metabolites
peak_data_scale_top50 <- peak_data_scale %>%
  select(PARENT_SAMPLE_NAME, select_variables$name) 
  

# Create heatmap data
heatmap_data <- create_heatmap(peak_data_scale_top50 ,chem_data,meta_data, meta_analysis_variables,
                                Gender, GROUP_NAME, desc(TIME1)) # Arranges data frame for heatmap



 
# Filter heatmap data to males only
Filtered_heatmap <- filter_heatmap(heatmap_data,drop_vars = "Gender",
                                   Gender=="Male") # identify the strata you want 
                                                    # to visualize. 


# Heat map colors 
palette <- colorRampPalette(rev(brewer.pal(10, "RdBu")))(256)


# Values for heatmap
vals = Filtered_heatmap$heatmap_data_vals

meta = Filtered_heatmap$heatmap_meta_data


# Create heatmap and save. 
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA)


# Save heatmap
pheatmap(vals, cluster_cols = F, cluster_rows = F, color = palette,
         annotation_col = meta, show_rownames = F, border_color = NA,filename = paste0("../Outputs/Plots/","HeatmapTop50MetabolitesStrat1.pdf"))






```

## Sub-pathway Analysis

Before diving in, lets load the matrices that we will need for the Sub-pathway Analysis
section. The three matrices we will need are:

1. Meta analysis data (meta_data): Contains all of the information about the samples. 

2. Scaled peak data (peak_data_scale):  The is the auto-scaled peak data derived
in the "Normalization and Standardization" section. 

3. Chemical annotation data (chem_data): Contains all of the information about the
metabolites, including which sub-pathway and super-pathway they belong to. 

```{r load-dataPrimary analysis}
# 1. Read meta-data
meta_data <- read.csv("../Data/Processed/meta_data.csv", check.names = F)

# 2. Read scaled peak data
peak_data_scale <- read.csv("../Data/Processed/peak_data_scale.csv", check.names = F)

# 3. Read chemical annotations
chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)
```

In the chemical annotation file, we will see that each metabolite is within a 
sub-pathway, and each sub-pathway is within a super-pathway. There are several 
metabolites within each sub-pathways and several sub-pathways within each 
Super-pathway. We can utilize an Analysis of variance (ANOVA) model to test for 
a difference in peak intensities between the treatment groups at the metabolite 
level, which is part of the Metabolon workflow. However, since multiple 
metabolites are within a sub-pathway, it is challenging to test if the treatment 
affected the peak data at the sub-pathway level. For this, we will be utilizing 
a combined fisher test. The combined Fisher test combines the p-values for each 
metabolite within a sub-pathway to test the same hypothesis at the sub-pathway 
level.

To test our hypothesis at the sub-pathway level, we first have to form our 
hypothesis at the metabolite level. For each metabolite, we test three models.

1.) Interaction: $log Peak = Treatment + Var2 + Treatment*Var2$

2.) Parallel: $log Peak = Treatment + Var2 $

3.) Single: $log Peak =  Treatment$

We are focusing only on the interaction term "Treatment*Var2" for this 
interaction model to test if there is a  significant interaction between our 
treatment and the second variable. The parallel model is testing if the second 
variable is explaining a significant amount of variance in the peak scaled 
intensity, and the treatment model is testing if the treatment explains a 
significant proportion of the variance in the scaled peak intensity for each 
metabolite. 

After running these three models for each metabolite, we will test at the 
sub-pathway level by combining the p-values of each model for each metabolite 
within the sub-pathway. We compute a chi-squared statistic to test at the 
sub-pathway level. For each model, we compute the chi-squared statistic by:

$$
\tilde{X} = -2\sum_{i=1}^k ln(p_i)
$$
where $k$ is the number of metabablites in the sub-pathway. We can get a p-value
from $P(X \geq\tilde{X})$, knowing that $\tilde{X}\sim \chi^2_{2k}$. 

#### Assumptions 
Since we are first testing each metabolite utilizing ANOVA, we make the following
assumptions for each metabolite,

* *Independence:* Each observation is independent of all other observations. 
Therefore, if you have collected multiple samples from the same subject then this
assumption may be violated. 

* *Normality:* The metabolite log-scaled intensities follow a normal distribution 
within each of the treatment groups. 

* *Homoscedasticity:* Equal variance between treatment groups. 

In addition to the assumptions in the ANOVA models at the metabolite level, the
Fisher's Combined probability places an independence assumptions for each metabolite
being tested within the sub-pathway. For example, when we test the interaction
model at the sub-pathway level, the p-values for interaction test are independent 
for each metabolite within the sub-pathway. 

### Combined Fisher Analysis

Before we run the combined fisher analysis, we need to create the analysis data,
which will contain all of the information required for the analysis. This data 
set includes the scaled peak data, and metadata all in one data frame. We will 
create this data and then save it to the Data/Processed folder. By default, this
data will save as "analysis.csv". You can change the saved name in the write.csv
function. Therefore, in the following chunk we:

1. Define the meta analysis variables to include in the analysis data set. 

2. Create the analysis data by merging the meta data, scaled peak data and the 
chemical annotation data. 

3. Save the analysis data to the "Data/Processed/" folder under the file name
"analysis_data"


```{r analysis data derivation}
# 1. Define meta analysis variables
meta_analysis_variables <- meta_analysis_variables <- c("PARENT_SAMPLE_NAME",
                             "GROUP_NAME",
                             "TIME1",
                             "Gender")


# 2. Create analysis data 
analysis_data <- peak_data_scale %>%
  pivot_longer(cols = -PARENT_SAMPLE_NAME,
               names_to = "CHEM_ID") %>%
  left_join(select(chem_data, CHEM_ID, CHEMICAL_NAME) %>%
              mutate(CHEM_ID= as.character(CHEM_ID)),
            by = "CHEM_ID") %>%
  left_join(select(meta_data, meta_analysis_variables),
            by = "PARENT_SAMPLE_NAME") %>%
  pivot_wider(id_cols = meta_analysis_variables,
              names_from = CHEMICAL_NAME, values_from = value)


# 3. Save the analysis data to Data/Processed folder
write.csv(analysis_data, file = "../Data/Processed/analysis_data.csv", row.names = F)
```



Now that we have our analysis data, we test the metabolomic data at the sub-pathway
level utilizing the `ancova_function` defined in the R script. This function takes
the following arguments:

* Anova data (anova_data): The data frame containing the combined meta, peak and 
chemical annotation data. 

* Meta analysis variables (meta_analysis_variables): The metadata variables which
includes the treatment groups and any other variables from the metadata. As a note
this should always include the "PARENT_SAMPLE_NAME" variable which is used to merge
the data frames. 

* Treatment only (treatment.only): This is a TRUE/FALSE. This should be set to TRUE
if you are only including a treatment variable and set to false if you are including
more than 1 variable. Setting this to TRUE will only run test the metabolites 
and the sub-pathways with the Single model. 

Treatment variable (treat_var): The name of the main treatment variable that needs
to be tested. 

* Additive variables (additive_vars): Defines which variables from the meta_analysis_variables
to include in the analysis. 

* Interaction variables (interaction_vars): Defines the interactions to include 
in the modeling. The set of variables to include in the interactions must be a 
subset of the variables from the additive_vars. An example of how this should look
look is $c("var1*var2")$. This will create interactions between var1
and var2.

*Note:* The ancova_function defined in the R script can only handle the modeling
with two variables. If additional variables are added, some modifications will need
to be made in the R script. 

To run this analysis, we utilize the following steps. 

1. Read in the analysis data and the chemical annotation data.

2. Define meta analysis variables to use in the modeling. 

3. Run the ancova_function with the arguments listed above

4. Save the pathway analysis results to the "Data/Results" folder with the name
"NonStratified_Full_ancova_results"

```{r run ancova}
# 1. Read in Analysis data and chemical annotation data
analysis_data <- read.csv("../Data/Processed/analysis_data.csv", check.names = F)

chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)

# 2. Define meta analysis variables
meta_analysis_variables <- meta_analysis_variables <- c("PARENT_SAMPLE_NAME",
                             "GROUP_NAME",
                             "TIME1",
                             "Gender")



# Run the ancova function
path_dat <- ancova_function(analysis_data,
                            chem_data = chem_data,
                                  meta_analysis_variables = meta_analysis_variables,
                                  treatment.only = F,
                                  additive_vars=c("GROUP_NAME","TIME1"),
                                interaction_vars =c("GROUP_NAME*TIME1"),
                                  treat_var = "GROUP_NAME")



# Save results
write.csv(path_dat,file = "../Data/Results/NonStratified_Full_ancova_results.csv", row.names = F)


```



### Stratified analysis

We may notice that we need to stratify our analysis if we believe the effects of
the model are different between the levels of the stratified variable. For 
example, we may notice that sex can change the effects of our treatment, and we 
may want to look at males and females separately. One way to do this is to subset
the data prior to running the ancova function.

In the following chunk you can specify a variable to stratify by. For each level
of the stratified variable, we:


1.) Subset the analysis data based on the stratified variable. 

2.) Run the ancova function with the subsetted data.

3.) Save the stratified data in "Data/Results" as "analysis_data_{Val}" where
{Val} is the value of the stratified variable. For example, if we stratified by sex 
then the data will be saved as "analysis_data_female for the females. 

4.) Save the results in "Data/Results" and the file will be saved as "Statified_{Val}_ancova_results.csv"
where {Val} is the value of the variable we stratified by. For example, if we stratified by sex, 
then the results for females would be saved as "Stratified_female_ancova_results.csv"

```{r stratifed analysis}

# 1. Read in analysis data
analysis_data <- read.csv("../Data/Processed/analysis_data.csv", check.names = F)

# define meta analysis variables
#meta_analysis_variables <- meta_analysis_variables <- c("PARENT_SAMPLE_NAME",
#                             "GROUP_NAME",
#                             "TIME1",
#                             "Gender")


# Defined stratified variable.
stratified_var = "Gender"

# Find unique values of this variable
uni_vals <- unique(analysis_data[,stratified_var])


# For each value perform the four steps listed above.
for(i in uni_vals){
  
  # 1. Subset analysis data
  strat_data = analysis_data[analysis_data[,stratified_var]==i,]
  
  
  # 2. Run ancova function on the subsetted data
  strat_path_results <- ancova_function(strat_data,
                                        chem_data = chem_data,
                                  meta_analysis_variables = meta_analysis_variables,
                                  treatment.only = F,
                                  additive_vars=c("GROUP_NAME","TIME1"),
                                interaction_vars =c("GROUP_NAME*TIME1"),
                                  treat_var = "GROUP_NAME")
  
  # 3 Save stratified data
  write.csv(strat_data, paste0("../Data/Processed/analysis_data_",i,".csv"), row.names = F)
  
  # 4 Save results
  write.csv(strat_path_results, paste0("../Data/Results/Stratified_",i,"_ancova_results.csv"), row.names = F)
  
  
}


```


#### Number of significant subpathways by model type

Once we have the Overall Analysis results, we will have tested all metabolites 
with an interaction, parallel and Single model. We then obtain a p-values through
the combined fisher analysis for each sub-pathway. In the next few chunks, we 
will summarize the results with a few different tables. The first summary will
show the number of significant sub-pathways for each model type. To do
this we:

1. Read in the results from the overall analysis. Currently we are reading in the 
results from one of the strata's in the stratified analysis. If you did not stratify 
the analysis you should change the path to the non stratified results. Additionally, 
if you stratified your analysis you may want to copy and paste this chunk to repeat
this table for each strata. 


2. Identify if the analysis only included a treatment variable. If you only included
a treatment variable, then treatment_only should be set to true. 

3. Structure the names of the levels for the model types in the table. 

4. Create the table data used to count the number of significant subpathways for each 
model type. 

5. Create the table by counting the number of sub-pathways for each model type. 

6. Display the table

7. Save the table in "Outputs/Tables" folder under the file name "NumberOfSigPathwaysByModelType_males".
You may want to change this name to reflect the strata you are summarizing. 

*Note:* If you only tested the sub-pathways for the Single model then only treatment 
is displayed. 

*Note:* The model type for each sub-pathway is determined hiearchically. For instance, 
if a sub-pathway shows a significant p-value for the interaction model and the parallel
model, then the sub-pathway model type is only "interaction". 

```{r table of significant subpathways}
# 1. Read in Results from the analysis step. 
path_data <- read.csv("../Data/Results/Stratified_Male_ancova_results.csv", check.names = F)


# 2. Treatment only: Change this to TRUE if only the treatment was tested. 
treatment_only = F


# 3. Structure levels
if(treatment_only){
  levs <- c("Single","None")
}
if(!treatment_only){
  levs =  c("Interaction", "Parallel", "Single", "None")
}


# 4. Create table data
table_data <- path_data %>%
  mutate(model = factor(model, levels = levs)) %>%
  select(sub_pathway, ends_with("_fisher"), model) %>%
  distinct()


# 5. Create table
 sig_subPaths <- count(table_data, model) %>%
                    arrange((model)) %>%
                    flextable() %>%
                   set_header_labels(model = "Model Type", n="Count")%>%
                   theme_vanilla() %>%
                   set_table_properties(layout = "autofit")
 
 
 # 6. Display table
 sig_subPaths

 
 # 7. Save table
 save_as_docx(sig_subPaths,path="../Outputs/Tables/NumberOfSigPathwaysByModelType_males.docx")
 
 
 # 8. Save table_data
 write.csv(table_data,"../Outputs/Tables/Table_data.csv", row.names = F)
```

#### Proportion of significant subpathways within super pathways:

Additionally, we can summarize the results by looking at the number of significant
sub-pathways within a super-pathway for each model type. This summary is done by:

1. Read in the table data and chem data.

2. Structuring the model levels.

3. Create the table data used to count the number of significant sub-pathways for each 
model type. 

4. Formulating the superpath data by:
  
  - Filtering all subpathways that did not have a significant subpathway. 
  
  - Join the results with the chemical annoation data to group by the superpathway. 
  
  - Summarize the percent of subpathways within superpathways that are significant. 
  
5. Display table 

6. Save table in the "Outputs/Tables" folder under the name "SigSuperPathwayPecentages"


*Note:* The model type for each sub-pathway is determined hierarchically. For instance, 
if a sub-pathway shows a significant p-value for the interaction model and the parallel
model, then the sub-pathway model type is only "interaction". 
```{r proportion in super pathways}
# 1. Read in table data and chem data 
table_data <- read.csv("../Outputs/Tables/Table_data.csv")
chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)


# 2. Structure Cases
if(sum(grepl("interaction",names(table_data))==0)){
  cases <- expression( case_when(model == "Single" ~ single_fisher))
}

if(sum(grepl("interaction",names(table_data)))>0){
  cases <- expression( case_when(model == "Interaction" ~ interaction_fisher,
                                        model == "Parallel" ~ parallel_fisher,
                                        model == "Single" ~ single_fisher))
}


# 3. Formulate the Superpathway results table.
superPath <- table_data %>% 
                filter(model != "None") %>%
                mutate(pval = eval(cases)) %>%
                select(sub_pathway, pval) %>%
                right_join(chem_data %>% select(SUPER_PATHWAY, SUB_PATHWAY) %>% distinct(),
                          by = c("sub_pathway" = "SUB_PATHWAY"))  %>%
                filter(!is.na(sub_pathway)) %>%
                mutate(sig = ifelse(is.na(pval), 0, 1)) %>%
                group_by(SUPER_PATHWAY) %>%
                summarise(percent_significant = round(mean(sig) * 100, 2),
                          number_significant = sum(sig),
                          pathway_count = n()) %>%
                ungroup() %>% arrange(-percent_significant) %>%
                transmute(SUPER_PATHWAY, percent_significant = paste0(number_significant, " / ", pathway_count, 
                                                                      " (", percent_significant, "%)")) %>%
                flextable() %>%
                set_header_labels(SUPER_PATHWAY="Super Pathway", percent_significant = "Percent Significant") %>%
                set_table_properties(layout = "autofit") %>%
                theme_vanilla()


# 4. Display table
superPath


# 5. Save table
save_as_docx(superPath,path = "../Outputs/Tables/SigSuperPathwayPecentages.docx")
```

#### List of subpathways that were found to have significant Fisher method p-values.

We may also be interested in obtaining a table which contains all of the sub-pathways
that had a significant model regardless of model type. To do this we create a table
in the following steps. 

1. Read in the table data

2. Structure model type levels.

3. Create pathway data that contains the sub-pathway, model type and the p-value 
for the model type with the the pathways that were non-significant filtered out. 

4. Format table

5. Display table

6. Save table in the "Outputs/Tables" folder with the name "SigSubpathwayTable"

```{r table of counts for each subpathway}
# 1. Read in table data
table_data <- read.csv("../Outputs/Tables/Table_data.csv")


# 2. Structure model type levels 
if(sum(grepl("interaction",names(table_data))==0)){
  cases <- expression( case_when(model == "Single" ~ single_fisher))
}

if(sum(grepl("interaction",names(table_data)))>0){
  cases <- expression( case_when(model == "Interaction" ~ interaction_fisher,
                                        model == "Parallel" ~ parallel_fisher,
                                        model == "Single" ~ single_fisher))
}


# 3. Create pathway table
pathway_table <- table_data %>% 
  filter(model != "None") %>%
  mutate(pval = eval(cases)) %>%
  select(sub_pathway, model, pval) %>%
  arrange(model, pval)


# 5. Format table
path_tab = pathway_table %>%
  flextable() %>%
  set_header_labels(sub_pathway="Subpathway", model = "Model Type", pval = "P-value") %>%
  set_table_properties(layout = "autofit") %>%
  colformat_double(digits = 3) %>%
  theme_vanilla()


# Display table
path_tab


# Save table
save_as_docx(path = "../Outputs/Tables/SigSubpathwayTable.docx")
```

For the subpathways with a significant model, it can also be useful to look at 
the results for the significant model for each metabolite within the significant
subpathway. In the following chunk we look at the metabolites within the top 2 
most significant sub-pathways for the interaction model and the parallel model. 
In the following chunk we:

1. Read in the table data and results from the overall analysis.

2. Structure model type levels.


3. Create the "top_pathway_table" which filters the table data to only focus on
metabolites with a significant model. Orders the data by model type and pvalue and then
selects the top five sub-pathways. Finally, we join this with the results from the
overall model to add metabolites and p-values for each of the top five significant
sub-pathways. 

4. For the top two sub-pathways with a significant interaction and parallel model
we do the following steps:

  - Select the top two sub-pathways. If you want  specific sub-pathways you can
  provide a list of the interested sub-pathways by changing inter_paths and par_paths to 
  c("subpathway_1", "subpathway_2", ...).
  
  - Filter the "top_pathway_table" on each of the selected sub-pathways. 
  
  - Display the metabolite name and p-value for the model
  
  - Each table will be displayed and saved to the "Outputs/Tables" folder under 
  the name metabolitesInTop{{model type}}Pathway.
  
```{r table of counts for each subpathway top five}
# 1. Read in table data and results from overall analysis. 
table_data <- read.csv("../Outputs/Tables/Table_data.csv")

path_data <- read.csv("../Data/Results/Stratified_Male_ancova_results.csv", check.names = F)


# 2. Structure model type levels 
if(sum(grepl("interaction",names(table_data))==0)){
  cases <- expression( case_when(model == "Single" ~ single_fisher))
}

if(sum(grepl("interaction",names(table_data)))>0){
  cases <- expression( case_when(model == "Interaction" ~ interaction_fisher,
                                        model == "Parallel" ~ parallel_fisher,
                                        model == "Single" ~ single_fisher))
}

# 3. Create top pathways table
top_pathway_table <- table_data %>% 
  filter(model != "None") %>%
  mutate(pval = eval(cases)) %>%
  select(sub_pathway, model, pval) %>%
  arrange(model, pval) %>%
  group_by(model) %>% slice(1:5) %>% ungroup() %>%
  left_join(select(path_data, sub_pathway, chem_name, ends_with("_pval")),
            by = "sub_pathway") %>%
  distinct()


# 4. Create and save tables

# Get top two pathways
inter_paths = unique(top_pathway_table[top_pathway_table$model=="Interaction",]$sub_pathway)[1:2]

# Top pathway (Display and Save)
top_pathway_table %>% 
  filter(sub_pathway == inter_paths[1]) %>%
  select(chem_name, interaction_pval) %>%
  arrange(interaction_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", interaction_pval="Interaction P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(inter_paths[1]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3)

top_pathway_table %>% 
  filter(sub_pathway == inter_paths[1]) %>%
  select(chem_name, interaction_pval) %>%
  arrange(interaction_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", interaction_pval="Interaction P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(inter_paths[1]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3) %>%
  save_as_docx(path = "../Outputs/Tables/metabolitesInTop1InteractionPathway.docx")


# Second Pathway
top_pathway_table %>% 
  filter(sub_pathway == inter_paths[2]) %>%
  select(chem_name, interaction_pval) %>%
  arrange(interaction_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", interaction_pval="Interaction P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(inter_paths[2]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3)

top_pathway_table %>% 
  filter(sub_pathway == inter_paths[2]) %>%
  select(chem_name, interaction_pval) %>%
  arrange(interaction_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", interaction_pval="Interaction P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(inter_paths[1]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3) %>%
  save_as_docx(path = "../Outputs/Tables/metabolitesInTop2InteractionPathway.docx")


# parallel --------- Top 2 paralell

par_paths = unique(top_pathway_table[top_pathway_table$model=="Parallel",]$sub_pathway)[1:2]

# Top 1 Parallel
top_pathway_table %>% 
  filter(sub_pathway == par_paths[1]) %>%
  select(chem_name, parallel_pval) %>%
  arrange(parallel_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", parallel_pval="Parallel P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(par_paths[1]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3)

top_pathway_table %>% 
  filter(sub_pathway == par_paths[1]) %>%
  select(chem_name, parallel_pval) %>%
  arrange(parallel_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", interaction_pval="Parallel P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(par_paths[1]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3) %>%
  save_as_docx(path = "../Outputs/Tables/metabolitesInTop1ParallelPathway.docx")


# Top 2 parallel

top_pathway_table %>% 
  filter(sub_pathway == par_paths[2]) %>%
  select(chem_name, parallel_pval) %>%
  arrange(parallel_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", parallel_pval="Parallel P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(par_paths[2]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3)

top_pathway_table %>% 
  filter(sub_pathway == par_paths[2]) %>%
  select(chem_name, parallel_pval) %>%
  arrange(parallel_pval) %>%
  flextable() %>%
  set_header_labels(chem_name="Metabolite", interaction_pval="Parallel P-Value") %>%
  set_table_properties(layout = "autofit") %>%
  add_header_row(values = c(par_paths[2]),colwidths = 2) %>%
  theme_vanilla() %>%
  align(i=1,align = "center", part = "header") %>%
  colformat_double(digits = 3) %>%
  save_as_docx(path = "../Outputs/Tables/metabolitesInTop2ParallelPathway.docx")



```


Finally, we would like to save a csv file with all of the significant sub-pathways.

1. Read in the table data

2. Structure the model levels

3. Filter table only on the significant subpathways

4. Save the file in the "Outputs/Tables" folder under the name "signifiant_subpathways". 
```{r save significant pathways in csv}
# 1. Read in the table data
table_data <- read.csv("../Outputs/Tables/Table_data.csv")


# 2. Structure the cases
if(sum(grepl("interaction",names(table_data))==0)){
  cases <- expression( case_when(model == "Single" ~ single_fisher))
}

if(sum(grepl("interaction",names(table_data)))>0){
  cases <- expression( case_when(model == "Interaction" ~ interaction_fisher,
                                        model == "Parallel" ~ parallel_fisher,
                                        model == "Single" ~ single_fisher))
}


# 3. Filter table only on the significant subpathways
sig_data <- table_data %>%
  filter(model != "None") %>%
  mutate(combined_fisher_p_val = eval(cases)) %>%
  select(sub_pathway, model, combined_fisher_p_val) %>%
  arrange(sub_pathway)


# 4. Save the file
write.csv(sig_data, file = "../Outputs/Tables/significant_sub_pathways.csv")

```


## Pairwise Compairisons

In the above analysis, we tested the interaction, parallel, and treatment models
for all sub-pathways. In this step, we tested if the sub-pathways were affected 
by any of the treatment groups. By first testing for all treatment groups, we can
preserve the family-wise error rate of 0.05. However, we may be interested in 
specific pairwise comparisons, and running multiple tests for each comparison of
interest increases our chances of making a type 1 error. If we subset the data to
only focus on the subpathways with a significant p-value when testing all 
treatment groups, then we can help preserve the family-wise error rate at 0.05.

1.) Reading in the data used in the analysis above. You will need to make changes
to the path to reflect the analysis data you want to use. For example the stratified
or non stratified analysis data. 

2.) Reading in the results from the analysis above. For example the stratified
or non stratified results.

3.) Sub-setting the analysis data to only include the sub-pathways that were found 
to be significant. This first step is to find the significant sub-pathways. By default
we are first going to look at the sub-pathways that had a significant interaction model. 

**Note:** Our code is currently set up to look at the results withing the male stratum
from the stratified analyse. You will want to change the file names to reflect the
names of the files generated from the results and analysis data above.

```{r get significant sub pathways}
# 1. Read in data from the analysis above
analysis_data <- read.csv("../Data/Processed/analysis_data_Male.csv", check.names = F)


# 2. Read in results from analysis above. 
path_data <- read.csv("../Data/Results/Stratified_Male_ancova_results.csv", check.names = F)


# 3. Filter data on pathways that showed a significant interaction.
significant_subpathways <- filter(path_data, model == "Interaction", !is.na(interaction_pval))

new_analysis_data <- analysis_data %>%
  select(PARENT_SAMPLE_NAME, GROUP_NAME, TIME1, Gender, significant_subpathways$chem_name)
  
```


The new analysis data will only include the sub-pathways with a significant 
interaction model. Now we want to look at all pairwise comparisons of our 
treatment variable. In the following chuck we are:

1.) Defining all of the meta data variables to include in the ancova_function. These
values can stay the same throughout the whole pipeline. 

2.) Defining all pairwise groups of the treatment variable. This is done with 
the get_compairisons function in dplyr. To make changes to this code you will need
to change the variable name within this function.


For each pair of values, we will add an additional argument to the ancova function
that will subset the data so that the analysis is only focused on the specified 
comparison. Within the for loop we are:

3.) Runnning the pairwise analysis. Any changes that are made to the arguments
of the anocova_function will also need to be made in the for loop. 

4.) Saving the pairwise results to Data/Results/Pairwise_Compairisons with the
default file name, "PairwiseCompairison_{var1}_and_{var2} where {var1} and {var2}
are the names of the values being compared.

Given the steps above, you will need to change the variable name in step 2 and
the variable name inside the for loop in step 3. 


```{r run ancova analysis on significant pathways}

# 1. Define all meta data variables to use in the analysis. 
meta_analysis_variables <- meta_analysis_variables <- c("PARENT_SAMPLE_NAME",
                              "GROUP_NAME",
                              "TIME1",
                              "Gender")



# 2. Define all pairwise groups of the treatment variable
pairs <- new_analysis_data %>%
  get_comparisons("GROUP_NAME") # Change argument as needed




# Run for loop for all of the pairwise compairsons
for (i in 1:length(pairs)){
  
  # 3. Run pairwise ancova_function
  pair <- pairs[[i]]
  
  
  pairwise_path_results <- ancova_function(new_analysis_data,
                                           chem_data = chem_data,
                                                  meta_analysis_variables = meta_analysis_variables,
                                                  treatment.only = F,
                                                  additive_vars=c("GROUP_NAME","TIME1"),
                                                  interaction_vars =c("GROUP_NAME*TIME1"),
                                                  treat_var = "GROUP_NAME", 
                                                  GROUP_NAME %in% pair) # Pairwise comparison 
                                                                        # Change variable name 
                                                                        # To a name within the meta data
  # 4. Save the pairwise comparison.
  write.csv(pairwise_path_results, paste0("../Data/Results/Pairwise_Compairisons/PairwiseCompairison_",
                                          pair[1], "_and_",pair[2],".csv"), row.names = FALSE)
  
}
```

In the chunk above, we tested each sub-pathway for several pairwise comparisons.
We now want to see which models were significant at the sub-pathway level for 
each comparison and which models were significant. The following code:

1.) Finds all of the saved pairwise results from the previous step.

2.) Gets the name of each pairwise comparison based on the file name.

3.) Create an initial table for the first comparison showing the number of 
sub-pathways that were significant for each model type. 

4.) Run a for loop for each pairwise comparison which creates the save table and
merges the number of sub-pathways that were significant for each model into one table.

5.) Format table and save. 

```{r pairwise table}

# 1. Find saved pairwise results
pairwise_files <- list.files("../Data/Results/Pairwise_Compairisons/")


# 2. Get the name of the each compairison based on the file name. 
pairs <- sub("PairwiseCompairison_","",sub(".csv","",pairwise_files))

# 3. Create initial table for first pairwise compairison
tablePair <- read.csv(paste0("../Data/Results/Pairwise_Compairisons/",pairwise_files[1]), check.names = F) %>% 
  select(sub_pathway, ends_with("_fisher"), model) %>%
  distinct() %>% 
  count(model, name = pairs[1]) 


# 4. For loop merging pairwise compairison results. 
for(i in 2:length(pairwise_files)){
  
  tablePair <- tablePair %>%
    full_join(read.csv(paste0("../Data/Results/Pairwise_Compairisons/",pairwise_files[i]), check.names = F) %>%
              select(sub_pathway, ends_with("_fisher"), model) %>%
              distinct() %>%
              count(model, name= pairs[i]),
            by = "model") 
  
}

# 5. Display and save table
tablePair %>%
  mutate(model = factor(model, levels = c("Interaction", "Parallel", "Single", "None")),
         across(-model, ~ifelse(is.na(.), 0, .))) %>%
  arrange((model)) %>%
  flextable() %>%
  set_header_labels(model="Model Type") %>%
  set_table_properties(layout = "autofit") %>%
  theme_vanilla()

tablePair %>%
  mutate(model = factor(model, levels = c("Interaction", "Parallel", "Single", "None")),
         across(-model, ~ifelse(is.na(.), 0, .))) %>%
  arrange((model)) %>%
  flextable() %>%
  set_header_labels(model="Model Type") %>%
  set_table_properties(layout = "autofit") %>%
  theme_vanilla() %>%
  save_as_docx(path = "../Outputs/Tables/SigPathsByModelType_pairwise.docx")

```


## Comparing all pairwise models

After running all of the pairwise comparisons and adding them to the 
Data/Results/Pairwise_Comparisons folder, we may want to look at the Fisher 
p-value for a specific model type for each pairwise comparison. For example, we 
may want to look at the interaction Fisher p-value across all comparisons for 
each sub-pathway.

In the following chunk we use a function called strat1_path_comparisons from the
.R file in the R script folder. This function takes the following arguments:

1. Path: A path to the folder storing the results for all pairwise comparisons.

2. Model: The column of the results that you would like to compare. For example,
we could compare interaction_fisher, parallel_fisher, or single_fisher across 
all pairwise comparisons.

The result will be a table with each row representing different sub pathways and
each column is the fisher p-value for the specified model. 



```{r pairwise comparisons}
# 1. Provide path
path = "../Data/Results/Pairwise_Compairisons/"


# 2. Create table compairing specified pvalues
strat1_path_comparisons <- sub_path_pairwise_comparison(path = path, interaction_fisher)
```


#### Comparison 1

Utilizing the table above, we can summarize our results in several ways. The next few chunks will be an example of how we can summarize the results. First, let's look at the names of the headers in the stratified pathways comparisons table we created above.

```{r names of strat table}
# Names of Strat1_path_comparisons.
names(strat1_path_comparisons)[-1]

```

Above are the names of the headers in the table comparing all pairwise comparisons
for a specific model. In the first table, we are creating a table that only
contains sub-pathways that show a significant p-value in the first comparison. 
To do this, will:

1. Filter the table so that only sub-pathways that were significant in the first 
comparison are displayed. To do this you will need to edit the code so that the 
correct comparison is used.

2. Select the sub_pathway and the specific comparison column. To do this you will 
need to edit the code so that the correct comparison is used. 

3. Rename to displayed header name to the desired header.

```{r compairison 1}
# Display sig. pathways for Comparison 1
comp1 <-strat1_path_comparisons %>%
  # 1. Filter on subpathways on the first comparison.
  filter(`1902_and_Control_interaction_fisher` < 0.05) %>%
  # 2. Select subpathway and comparison. 
  select(sub_pathway, `1902_and_Control_interaction_fisher`  ) %>%
  arrange(sub_pathway) %>%
  flextable() %>%
  # 3. Change header name to something more readable. 
  set_header_labels(sub_pathway="Sub Pathway", `1902_and_Control_interaction_fisher` ="Combined Fisher Prob P-val") %>%
  colformat_double(digits = 3) %>%
  set_table_properties(layout = "autofit") %>%
  theme_vanilla()
  

# Display table
comp1
  
# Save table
comp1 %>%
  save_as_docx(path = "../Outputs/Tables/SigPathwaysPairwiseComparison1.docx")

```

#### Comparison 2
Now that we know how many sub-pathways are significant for the first comparison, 
we might want to check which sub-pathways were not significant in the first 
sub-pathway but were significant in the second comparison. Similar to the table above, we are
going to:


1. Filter the table so that only sub-pathways that were not significant in the first 
comparison and significant in the second comparison are displayed. To do this you
will need to edit the code so that the correct comparisons are used.

2. Select the sub_pathway and the specific comparison column.To do this you will
need to edit the code so that the correct comparison is used. 

3. Rename to displayed header name to the desired label. 


```{r comparison 2}
# Table displays non-significant comparison 1 and significant comparison 2
comp2 <- strat1_path_comparisons %>%
  # 1. Filter on subpathways on the first comparison.
  filter(`1902_and_Control_interaction_fisher`  >= 0.05, `1902_and_WT_interaction_fisher`  < 0.05) %>%
  # 2. Select subpathway and comparison. 
  select(sub_pathway, `1902_and_WT_interaction_fisher` ) %>%
  arrange(sub_pathway) %>%
  flextable() %>%
  # 3. Change header name to something more readable.
  set_header_labels(sub_pathway="Sub Pathway", `1902_and_WT_interaction_fisher` ="Combined Fisher Prob P-val") %>%
  colformat_double(digits = 3) %>%
  set_table_properties(layout = "autofit") %>%
  theme_vanilla()
  
# Display table
comp2
  

# Save table
comp2 %>%
  save_as_docx(path = "../Outputs/Tables/SigPathwaysPairwiseComparison2.docx" )
  


```

#### Comparison 3

Now that we know how many sub-pathways are significant for the first two comparisons, 
we might want to check which sub-pathways were significant in the third sub-pathway
but not the first two. This pattern can go on for as many comparisons as necessary.
Additionally, we can augment the "filter" arguments to display the table as desired.
Similar to the table above, we are going to:


1. Filter the table so that only sub-pathways that were not significant in the first 
comparison and significant in the second comparison are displayed. To do this you
will need to edit the code so that the correct comparisons are used.

2. Select the sub_pathway and the specific comparison column.To do this you will need to edit the code so that the 
correct comparison is used. 

3. Rename to displayed header name to the desired label. 

```{r}
# Table displays non-significant comparison 1 and 2 and significant comparison 3
comp3 <- strat1_path_comparisons %>%
  # 1. Filter on subpathways on the first comparison.
  filter(`1902_and_Control_interaction_fisher`  >= 0.05, `1902_and_WT_interaction_fisher` >= 0.05,
          Control_and_WT_interaction_fisher < 0.05) %>%
   # 2. Select subpathway and comparison
  select(sub_pathway, Control_and_WT_interaction_fisher) %>%
  arrange(sub_pathway) %>%
  flextable() %>%
  # 3. Change header name to something more readable.
  set_header_labels(sub_pathway="Sub Pathway", Control_and_WT_interaction_fisher="Combined Fisher Prob P-val") %>%
  colformat_double(digits = 3) %>%
  set_table_properties(layout = "autofit") %>%
  theme_vanilla()
  

# Display table
comp3

# Save table
comp3 %>%
  save_as_docx(path = "../Outputs/Tables/SigPathwaysPairwiseComparison3.docx")


```




#### All comparisons
We may want to look at specific sub-pathways for each comparison and all models. 
In the chunk below we use the sub_path_pairwise function defined in the R scripts
folder. This function takes two arguments. 

* path: The path the pairwise comparison results folder.

* ... This is a masking parameter that allows you to filter the sub_pathway to the
specific subpathway of interest. 

In the example below we:

1. Provide a path to the Pairwise comparison results.

2. Create the table with all fisher p-values for the selected subpathways. 

3. Displays the table

4. Saves table in ../Outputs/Tables/ with the file name SubpathwaysFisherProbs_allModels.


```{r subpathway all comparisons}

# 1. Establish path to pairwise results
path =   "../Data/Results/Pairwise_Compairisons/"

  

# 2. Create the table with all fisher p-values for the selected subpathways.
tab_all <- sub_path_pairwise_comparison_all(path =path, sub_pathway %in%  c("Vitamin B6 Metabolism", "Creatine Metabolism", "Pentose Metabolism"))

#2. Create the table with all fisher p-values for the selected subpathways
tab_all

# 4. Saves table in ../Outputs/Tables/ with the file name SubpathwaysFisherProbs_allModels.
tab_all %>%
  save_as_docx(path = "../Outputs/Tables/SubpathwaysFisherProbs_allModels.docx")


```


## Plots
Visualizations of the data can help us see the underlying trends. One useful visualization
is boxplots and line graphs. 

### Boxplots

Suppose there is a specific subpathway that we would like to visualize. In that 
case, we can compare the treatment groups for each metabolite within that 
sub-pathway using the subpathway_box plots function defined in the R script. T
his function takes the following arguments.

* Analysis data (analysis_data): This is the analysis data used for the modeling
in the primary analysis. If you followed this pipeline all the way through, the 
analysis data is saved in the Data/Processed folder. 

* Chemical annotation data (chem_data): The chemical annotation data. This argument
is necessary to link the columns of the analysis data to the sub
pathways. If you followed this pipeline all the way through, the analysis data is
saved in the Data/Processed folder. 

* Pathway (pathway): This is the name of the sub-pathway of interest. 

* X: This the the name of the variable in the meta data that is used for the 
X axis of the box plots. 

* Group By (groupBy): This is a grouping variable. As a recommendation the treatment
groups should be used in the groupBy argument as this will provide a different color
for each of the treatments making it easier to identify. 

* ... : At the end of this function you can provide additional arguments for filtering
the analysis data. In the examples below we are filtering so the box plots only 
focus on males and only two treatment groups. 

The output of the subpathway_boxplot function is box plots for each of the 
metabolites within the specified subpathway. The labels for the box plots are 
default based on the data created inside the subpathway_boxplots function. Since
these box plots utilize ggplot, we can customize all labels using the. [labs](https://ggplot2.tidyverse.org/reference/labs.html) function
from ggplot. 

#### Box plots example
Here we look at the boxplots for each metabolite within a specified subpathway utilizing
the subpathway_boxplots function. In the following chunk we:

1. Read in the analysis data

2. Read in the chemical annotation data

3. Specify the sub-pathway of interest

4. Run the subpathway_boxplots function

5. Customize labels for the generated box plot. 

6. Save the final boxplots to the Outputs/Figures folder with the name
"Metabolite_boxplots_{{subpathway}}" where {{subpathway}} is the name of the 
specified sub-pathway. 


```{r boxplots, fig.width = 10, fig.height=10}

# 1. Read in analysis data
analysis_data <- read.csv("../Data/Processed/analysis_data_Male.csv", check.names = F)

# 2. Reach in chemical annotation data
chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)

# 3. Specify subpathway of interest. 
subpath = "Gamma-glutamyl Amino Acid"


# 4. Run the subpathway_boxplots function
box_1 <- subpathway_boxplots(analysis_data =  analysis_data,
                chem_data = chem_data,
                subpathway = subpath,
                X = TIME1,
                groupBy = GROUP_NAME,
                # Additional analysis data filtering options. 
                Gender == "Male", GROUP_NAME != "Control")


# 5. Customize labels for the boxplots
box_1 +
  labs(x = "Time", y="Scaled Intensity", color="Treatment Group",
       title = "Metabolite boxplots for the Gamma-glutamyl Amino Acid Subpathway")


# 6. Save boxplots. You can change the file type for example change pdf to png. 
ggsave(filename = paste0("../Outputs/Plots/Metabolite_boxplots_",subpath,".pdf"))

```



### Line plots

Another useful visualization is line plots. These plots are useful for seeing 
trends between an ordered variable and the scaled intensity. Line plots can be 
useful for analyzing trends across the different treatment groups. Like the box 
plots, we are interested in seeing trends for all metabolites within  a sub-pathway.
To do this, we can utilize the "subpathway_lineplots" function defined in the R 
scripts. This function takes the following arguments.

* Analysis data (analysis_data): This is the analysis data used for the modeling
in the primary analysis. If you followed this pipeline all the way through, the 
analysis data is saved in the Data/Processed folder. 

* Chemical annotation data (chem_data): The chemical annotation data. This argument
is necessary to link the columns of the analysis data to the subpathways. If you
followed this pipeline all the way through, the analysis data is saved in the Data/Processed folder. 

* Pathway (pathway): This is the name of the subpathway of interest. 

* X: This the  name of the variable in the metadata that is used for the 
X axis of the line plots. 

* Group By (groupBy): This is a grouping variable. As a recommendation the treatment
groups should be used in the groupBy argument as this will provide a different color
for each of the treatments making it easier to identify. 



#### Line plot example

Below is a demonstration of utilizing the line plots to see trends between 
treatment groups at the metabolite level. In the chunk below we:

1. Read in the analysis data

2. Read in the chemical annotation data

3. Prep data for line plots. To produce the lines we must provide a numeric
variable for the x-axis. If you are looking at trends across an ordinal categorical
variable you will need to convert the variable to be numeric. For example, if we 
want to see the trend in time between treatment groups, where time takes the ordinal
values of "Pre-symptomatic", "onset of symptoms", and "end of symptoms", then we
will need to assign these names to the values of 1,2, and 3 respectively. 
Additionally, we will need to make any other desired adjustments to our data at 
this step. 


4. Define the sub-pathway of interest. 

5. Run the "subpathway_lineplots function"

6. Edit the aesthetics of plots. Similar to the box plots, we are using the labs
function to edit the labels for the plots. Additionally, we are utilizing the 
scale_x_continuous function at add labels to the tick marks for the plot that 
correspond with the original values. 

7. Save the plots in the folder "Outputs/Plots" under the file name 
"linePlots_{{subpathway}} where {{subpathway}} is the name of the subpathway 
defined in step 4.


```{r line-plots}
# 1. Read in analysis data
analysis_data <- read.csv("../Data/Processed/analysis_data.csv", check.names = F)

# 2. Reach in chemical annotation data
chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)


# 3. Prep data for line plots
analysis_data <- analysis_data %>%
  filter(Gender=="Male") %>%
  mutate(TIME1 = case_when(TIME1== "PreSymp" ~ 1,
                           TIME1 =="Onset" ~ 2,
                       TIME1 == "End" ~ 3))



# 4. Define subpathway of interest. 
subpathway = "Glycerolipid Metabolism"


# 5. Run subpathway_line plots function.
line_plots <- subpathway_lineplots(analysis_data = analysis_data,
                     chem_data = chem_data,
                     subpathway = subpathway,
                     X = TIME1,
                     groupBy = GROUP_NAME)



# 6. Edit asthetics of plots
line_plots +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("PreSymp", "Onset", "End")) +
  labs(x= "Time",y="Scaled Intensity",color = "Treatment Group")


# 7. Save plots 
ggsave(filename = paste0("../Outputs/Plots/LinePlots_",subpathway,".pdf"))

```


#### Stratified line plots

We may also want to stratify the line plots by a specific condition. In the example
below we stratify the line plots by gender in 10 steps. 

1. Read in the analysis data

2. Read in the chemical annotation data

3. Define the subpathway of interest. 

4. Prep data for line plots and filter on specific conditions in the data. In this
example we are filtering the data on males. Additionally, the same processing steps 
from above need to be completed.

5. Filter analysis data on second condition following the same instructions from
step 3. In this example we are filtering the data on females. 

6. Run the subpathway_lineplots function for males

7. Run the subpathway_lineplots function for females

8. Use the ggarrange function to combine these two plots into one. 

9. Save the plot in Outputs/Plots folder with the LinePlots_{{subpathway}}_stratified.



```{r line-plots stratified}
# 1. Read in analysis data
analysis_data <- read.csv("../Data/Processed/analysis_data.csv", check.names = F)

# 2. Reach in chemical annotation data
chem_data <- read.csv("../Data/Processed/chem_data.csv", check.names = F)




# 3. Define subpathway of interest. 
subpathway = "Glycerolipid Metabolism"


# 4. Prep data for line plots
analysis_data_male <- analysis_data %>%
  filter(Gender=="Male") %>%
  mutate(TIME1 = case_when(TIME1== "PreSymp" ~ 1,
                           TIME1 =="Onset" ~ 2,
                       TIME1 == "End" ~ 3))


# 5. Prep data for line plots
analysis_data_female <- analysis_data %>%
  filter(Gender=="Female") %>%
  mutate(TIME1 = case_when(TIME1== "PreSymp" ~ 1,
                           TIME1 =="Onset" ~ 2,
                       TIME1 == "End" ~ 3))




# 6. Create plot for males
males <- subpathway_lineplots(analysis_data = analysis_data_male,
                     chem_data = chem_data,
                     subpathway = subpathway,
                     X = TIME1,
                     groupBy = GROUP_NAME)+
    scale_x_continuous(breaks = c(1, 2, 3), labels = c("PreSymp", "Onset", "End")) +
    labs(x= "Time",y="Scaled Intensity",color = "Treatment Group")



# 7. Create plot for females
females <- subpathway_lineplots(analysis_data = analysis_data_male,
                     chem_data = chem_data,
                     subpathway = subpathway,
                     X = TIME1,
                     groupBy = GROUP_NAME)+
    scale_x_continuous(breaks = c(1, 2, 3), labels = c("PreSymp", "Onset", "End")) +
    labs(x= "Time",y="Scaled Intensity",color = "Treatment Group")


# 8. Display stratified plot
ggarrange(males,
  females,
  labels = c("Males", "Females"),
  nrow = 2,
  common.legend = T,
  vjust = 0
) 





# 9. Save plots 
ggsave(filename = paste0("../Outputs/Plots/LinePlots_",subpathway,"_stratified.pdf"))


```


