---
title: "MetabolomicsPipeline-vignette"
author: "Joel Parker"
date: "05/31/2024"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{MetabolomicsPipeline-vignette}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---


## Installation
You can install the MetabolomicsPipeline package directly from BioConductor 
using the following code. Note, this package is not yet available on 
BioConductor. 
```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("MetabolomicsPipeline", force = TRUE)
```

You can additionally install the current version which is hosted on github using 
the following code.

```{r eval=FALSE}
if (!requireNamespace("devtools", quietly=TRUE))
    install.packages("devtools")
    
 devtools::install_github("datalifecycle-ua/MetabolomicsPipeline", build_vignettes = TRUE)
```


Once the MetabolomicsPipeline package is installed, we can load it into the 
environment. Note we are also loading the "table1" package. This package
is not required by the MetabolomicPipeline, however, the table1 package contains
great functions for creating tables. Specifically, it is useful for showing the 
number of samples in each of our experimental groups. We recommend installing 
this package using install.packages("table1"). The "SummarizedExperiment" and
"ggplot2" packages are required to install the MetabolomicsPipeline package and
will being automatically installed (if not already installed) when installing
the MetabolomicsPipline package using the commands above. 


```{r setup}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 10
)

# Tables
library(table1)

# Load Metabolomics Pipeline
library(MetabolomicsPipeline)

library(SummarizedExperiment)

# Figures
library(ggplot2)
```


## Introduction

The purpose of the MetabolomicPipeline package is to streamline the analysis for targeted
metabolomics experiments. Targeted metabolomics is useful for hypothesis driven experimental designs, focusing the analysis on specific metabolites or pathways of interest. In this vignette we demonstrate how 
to use MetabolomicsPipeline package for:

* Data Processing

* Exploratory Analysis

* Subpathway Analysis

* Metabolite pairwise comparisons

* Subpathway box plots and line plots

### Comparison with other packages
Several R packages are available for metabolomics data analysis. The MetabolomicsPipeline package is designed to streamline workflows after metabolite identification and peak quantification. Unlike other tools, it enables hypothesis testing at the subpathway level by leveraging univariate ANOVA results at the metabolite level. It also provides a straightforward framework for performing pairwise comparisons between metabolites.

To our knowledge, MetabolomicsPipeline is the only R package that implements the combined Fisher probability test for detecting differences at the subpathway level. All outputs, including pairwise comparisons, are presented in publication-ready tables, facilitating a seamless transition from analysis to publication.

### Data Description

In this vignette, we will use data which consists of 86 samples (42
males, 44 females), three treatment groups, and the samples were taken
at three different time points.


### Metabolomics data as a SummarizedExperiment

Metabolomics experiments leverage multiple data tables for analysis.
The datasets needed for the downstream
analysis are:

1.) Sample metadata

2.) Chemical annotation

3.) Peak data (samples x rows).


### Data loading

The MetabolomicsPipeline package provides a convenient way to load each
of these datasets together as a [SummarizedExperiment](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html) using create_met_se(). In this chunk we load the
demo sample metadata, chemical annotation, and peak data into a SummarizedExperiment.

1. load the demo sample metadata, chemical annotation, and peak data into a
SummarizedExperiment.


2. Create a table of the sample distribution


```{r AnalysisDataCreation}
################################################################################
### Load Data ##################################################################
################################################################################

# if sample metadata, chemical annotation and peak data are stored in .xlsx
# you can use.

# dat <- load_met_excel(
  # path,
  # raw_sheet = "Peak Area Data",
  # chemical_sheet = "Chemical Annotation",
  # sample_meta = "Sample Meta Data",
  # normalized_peak = "Log Transformed Data",
  # sample_names = "PARENT_SAMPLE_NAME",
  # chemicalID = "CHEM_ID"
# )

# Get demo sample metadata
data("demoSampleMeta", package = "MetabolomicsPipeline")

# Get demo chemical annotation file
data("demoChemAnno", package = "MetabolomicsPipeline")

# Get demo peak data
data("demoPeak", package = "MetabolomicsPipeline")

dat <- createMetSe(chemical_annotation = demoChemAnno,
                     sample_metadata = demoSampleMeta,
                     peak_data = demoPeak,
                     chemical_id = "CHEM_ID",
                     sample_names = "PARENT_SAMPLE_NAME")


################################################################################
### Create Table 1 #############################################################
################################################################################
# Create table 1
tbl1 <- table1(~ GROUP_NAME + TIME1 | Gender,
  data = colData(dat)
)

# Display table 1
tbl1
```


## Data Processing
The MetabolomicsPipline package contains tools for processing the peak data
to prepare it for downstream analysis:

1.) Median standardization (medianStandardization())

2.) Minimum value imputation (minValImpute())

3.) Log transformation (logTransformation())

```{r Data Processing}
# Median standardization
dat <- medianStandardization(dat, assay = "peak")

# Min value imputation
dat <- minValImpute(dat, assay = "median_std") 

# log transformation
dat <- logTransformation(dat, assay = "min_impute")
```


## Exploratory Analysis

In data exploration, we use several methods to help us better understand
the underlying patterns in the data without using a formal hypothesis
test. In this pipeline, we are going to focus on two methods of data
exploration:

A.) Principal component analysis

B.) Heatmaps

#### Principal Component Analysis (PCA)

In general, Principal component analysis (PCA) reduces the number of
variables in a dataset while preserving as much information from the
data as possible. At a high level, PCA is constructed such that the
first principal component (PC) accounts for the largest amount of
variance within the data. The second PC accounts for
the largest remaining variance, and so on. Additionally, each
of the PCs produced by PCA is uncorrelated with 
the other principal components. PCA can allow us to visualize sources of
variation in the data. The metabolite_pca function will enable us to specify
a sample metadata variable to label the points in the plot. The metabolitePca 
function has three arguments:

* data: SummarizedExperiment with metabolomics experiment data.

* Assay: Name of the assay to be used for the pairwise analysis
    (default='normalized')

* meta_var: A metadata variable to color code the PCA plot by.


```{r ExploratoryAnalysis_PCA}
###############################################################################
### Run PCA ###################################################################
###############################################################################

# Define PCA label from metadata
meta_var <- "Gender"

# Run PCA
pca <- metabolitePca(dat,
    meta_var = meta_var
)


# Show PCA
pca
```

Suppose you notice a variable with clearly separated groups that is not a 
variable of interest. In that case, consider stratifying your downstream
analysis by the values of that variable. For example, we will stratify the 
downstream analysis by male/female in our vignette data. 

#### Heatmaps

For our heatmap, the
x-axis will be the samples, and the y-axis will be the metabolites. The
values determining the colors will be the log normalized peak values for each
metabolite in each observation. We can group the observations by the 
experimental conditions. Grouping the experimental conditions in a heatmap is 
another way of visualizing sources of variation within our data.

We can use the metaboliteHeatmap function to create the heatmaps, which 
requires the following arguments. 

* data: SummarizedExperiment with metabolomics experiment data.

* top_mets: The number of metabolites to include in the heatmap. The metabolites
    are chosen based on the metabolites with the highest variability. 

* group_vars: The variables to group the samples by.Note, if you want to order
the columns of the heatmap based on a group, you must include them in the 
group_vars. 
    

* caption: A title for the heatmap. If strat_var is used, the title will 
    automatically include the stratum with the tile.

* Assay: Which assay data to use for the heatmap (default="normalized").

* ... : You can add additional arguments to order the samples


In the chunk below, we create a PCA plot labeled by Gender. Then, we make three
heatmaps increasing by complexity. 

```{r ExploratoryAnalysis_heatmap}
################################################################################
### Run Heatmaps ###############################################################
################################################################################

# Heatmap with one group
metaboliteHeatmap(dat,
    top_mets = 50,
    group_vars = "GROUP_NAME",
    strat_var = NULL,
    caption = "Heatmap Arranged By Group",
    Assay = "normalized",
    GROUP_NAME
)




# Heatmap with two groups
metaboliteHeatmap(dat,
    top_mets = 50,
    group_vars = c("GROUP_NAME", "TIME1"),
    strat_var = NULL,
    caption = "Heatmap Arranged By Group and TIME",
    Assay = "normalized",
    GROUP_NAME, desc(TIME1)
)


# Heatmap with 2 group and stratified
metaboliteHeatmap(dat,
    top_mets = 50,
    group_vars = c("GROUP_NAME", "TIME1"),
    strat_var = "Gender",
    caption = "Heatmap Arranged By Group and TIME",
    Assay = "normalized",
    GROUP_NAME, desc(TIME1)
)


```

## Subpathway Analysis

In the chemical annotation file, we will see that each metabolite is
within a subpathway, and each subpathway is within a superpathway.
There are several metabolites within each subpathway and several
subpathways within each superpathway. We can utilize an Analysis of
variance (ANOVA) model to test for a difference in peak intensities
between the treatment groups at the metabolite level. However, since multiple metabolites are
within a subpathway, it is challenging to test if the treatment
affected the peak data at the subpathway level. For this, we 
utilize a combined Fisher probability test. The combined Fisher test
combines the p-values from independent tests to test the hypothesis for
an overall effect. The Combined Fisher Probability is helpful
for testing a model at the subpathway level based on the pvalues from 
the model at the metabolite level.

### Combined Fished Analysis

We will test at the subpathway level by combining the p-values for each
metabolite within the subpathway for each model. We use a combination
function given by $\tilde{X}$ which combines the pvalues, resulting in a
chi-squared test statistic.

$$
\tilde{X} = -2\sum_{i=1}^k ln(p_i)
$$
where $k$ is the number of metabolites in the subpathway. We can
get a p-value from $P(X \geq\tilde{X})$, knowing that
$\tilde{X}\sim \chi^2_{2k}$. You will notice that smaller p-values will
lead to a larger $\tilde{X}$.

##### Assumptions

Since we are first testing each metabolite utilizing ANOVA, we make the
following assumptions for each metabolite,

-   *Independence:* Each observation is independent of all other
    observations. Therefore, if you have collected multiple samples from
    the same subject then this assumption may be violated.

-   *Normality:* The metabolite log-scaled intensities follow a normal
    distribution within each of the treatment groups.

-   *Homoscedasticity:* Equal variance between treatment groups.

In addition to the assumptions in the ANOVA models at the metabolite
level, the Fisher's Combined probability places an independence
assumption between the metabolites within the subpathway. 

For more about the Combined Fisher Probability and other methods that
can address this problem, see:

Loughin, Thomas M. "A systematic comparison of methods for combining
p-values from independent tests." Computational statistics & data
analysis 47.3 (2004): 467-485.

#### Models

To test our hypothesis at the subpathway level, we first have to form
our hypothesis at the metabolite level. For each metabolite, we test
three models.

1.) Interaction: $log Peak = Treatment Group + Time + Treatment*Time$

2.) Parallel: $log Peak = Treatment Group + Time$

3.) Single: $log Peak = Treatment$

For the interaction model, we are focusing only on the interaction term
"Treatment\*Time" to test if there is a significant
interaction between our treatment and the time variable. The parallel
model is testing if the time variable is explaining a significant
amount of the metabolite variance with treatment included, and the treatment
model is testing if the treatment explains a significant proportion of the 
variance for each metabolite.  

We test at the subpathway level using the Combined Fisher Probability method to 
combine the p-values from each model for all metabolites within the subpathway. 

Model selection is based on the combined Fisher p-value. The interaction model is
selected if the interaction term in Model 1 is significant. If the blocking factor
(TIME1) is significant in Model 2, the parallel model is chosen. If only the 
treatment term is significant, the single-factor model is selected.

To run the subpathway analysis, we use the "subpathwayAnalysis"
function, which requires the following arguments.  


* data: SummarizedExperiment with metabolomics experiment data.

* treat_var: This is the name of the variable in the analysis data that is the 
    main variable of interest.

* block_var: This is the name of the blocking variable in the dataset. If the
    experimental design does not include a blocking variable, then the value of 
    block_var=NULL.

* strat_var: Variable to stratify the subpathway analysis by. This is set to 
    NULL by default and will not stratify the analysis unless specified.

* Assay: Name of the assay to be used for the pairwise analysis 
    (default='normalized')


#### Results Summaries

With the MetabolomicsPipeline package, we provide three different ways to
summarize the results from the subpathway analysis.

1.  Number of significant subpathways by model type (subpathByModel)

2.  Percentage of significant subpathways within superpathways 
    (subpathWithinSuperpath)
    
3. Subpathway results within a specified superpathway (subWithinSuperResults)

4.  Metabolite model results within a specified subpathway (metWithinSub)


**Note:** If any of the ANOVA models above have a warning or an error, the 
metabolite is excluded and a warning is printed with the message
about the warning or error. Metabolites with a warning or error are excluded in the 
subpathway analysis.

```{r SubpathwayAnalysis}
################################################################################
## Stratified Analysis #########################################################
################################################################################

# Stratified Analysis
stratified <- subpathwayAnalysis(dat,
    treat_var = "GROUP_NAME",
    block_var = "TIME1",
    strat_var = "Gender",
    Assay = "normalized"
)


################################################################################
### Results Plots ##############################################################
################################################################################

# 1. significant subpathways by model type
subpathByModel(stratified)

# 2. Percentage of signficant subpathways within superpathways
subpathWithinSuperpath(stratified)

# Show significant subpathways within superpathways
subWithinSuperResults(stratified, Superpath = "Amino Acid")

# 3. Metabolites within subpathway
tables <- metWithinSub(stratified,
    subpathway = "Lysine Metabolism"
)


### Females
tables$Female

### Males
tables$Male
```

## Pairwise Analysis

We can look at the pairwise comparisons for all
experimental groups at the metabolite level. We will use the metabolite_pairwise function within the
MetabolomicsPipeline package, which requires the following arguments: 


* data: SummarizedExperiment with metabolomics experiment data.

* Assay: Name of the assay to be used for the pairwise analysis 
    (default='normalized')

* mets: Chemical ID for the metabolites of interest. If NULL then the pairwise
    analysis is completed for all metabololites.

* form: This is a character string the resembles the right hand side of a simple
    linear regression model in R. For example form = "Group1 + Group2".

* strat_var: A variable in the analysis data to stratify the model by. If this 
    is specified, a list of results will be returned.

#### Log Fold-Change Heatmap

We will produce a heatmap of the log fold changes for the metabolites with a 
significant overall p-value (which tested if the treatment group means were 
equal under the null hypothesis). The
heatmap colors will only show if the log fold-change is greater than
log(2) or less than log(.5). Therefore, this heatmap will only focus on
comparisons with a fold change of two or greater. The met_est_heatmap function 
will produce an interactive heatmap using the results from the pairwise
analysis.  

#### P-Value Heatmap

Similar to the pairwise estimate heatmap, we will produce a heatmap where the 
heatmap will only include metabolites with a significant overall p-value, and 
the values in the heat map will only be colored if the pairwise comparison is 
significant. We use the metPHeatmap function to create an interactive p-value
heatmap. 


For both the log fold-change heatmap and the p-value heatmap, there is an option
to produce an interactive plot using plotly or to create a static heatmap using
pheatmap. To produce these heatmaps we use the following arguments:


*  results_data: Results data frame of the pairwise comparisons produced by
    \code{\link{metabolitePairwise}}.

*  data: A SummarizedExperiment containing the metabolomics experiment data.

* FC_cutoff: Numeric value for the fold change cutoff. Must be larger than 1. Default is 2.

* pv_cutoff: Numeric value for the p-value cutoff. Default is 0.05.

*  interactive: boolean (TRUE/FALSE) for whether or not the plot should be
    interactive. Use interactive=T to produce an interactive plot using 
    plotly. Use interactive=FALSE to produce a static heatmap using pheatmap.
    
* plotlyTitle: Title for the interactive heatmap

* ...: Additional arguments that can be passed to pheatmap. This is only for the
static heatmap. 

```{r PairwiseAnalysis}
################################################################################
#### Run Pairwise Comparisons ##################################################
################################################################################

strat_pairwise <- metabolitePairwise(dat,
    form = "GROUP_NAME*TIME1",
    strat_var = "Gender"
)


###############################################################################
## Create Estimate Heatmap #####################################################
################################################################################

metEstHeatmap(strat_pairwise$Female, dat,
              diff_cutoff = .7, pv_cutoff = 0.05,
    interactive = TRUE,
     SUB_PATHWAY = "SUB_PATHWAY",
    CHEMICAL_NAME = "CHEMICAL_NAME",
    plotlyTitle = "Estimate Heatmap",
    main = "Log fold change heatmap", show_rownames = FALSE
)


################################################################################
## Create P-value Heatmap ######################################################
################################################################################
# Female
metPHeatmap(strat_pairwise$Female, dat,
            pv_cutoff = 0.05,
    interactive = TRUE, show_rownames = FALSE,
    plotlyTitle = "Pvalue Heatmap",
    main = "Pvalue Heatmap"
)
```



## Boxplots and Lineplots

Visualizations of the data can help us see the underlying trends. Two
useful visualizations are boxplots and line plots, we will be using the
subpathwayBoxplots and subpathwayLineplots functions to create them. The main 
utility of these functions is it allows you for focus on the metabolites within 
a subpathway. For both functions, the arguments are:


* data: SummarizedExperiment with metabolomics experiment data.

* subpathway:	Character value of the subpathway of interest. This is case
    sensitive and must be in the chemical annotation file.

* block_var: This the the name of the variable in the meta data that is used for
    the X axis of the box plots. We recommend using the "block_var" from the 
    subpathway analysis.

* treat_var: This is a grouping variable. As a recommendation the treatment 
    groups should be used in the treat_var argument as this will provide a 
    different color for each of the treatments making it easier to identify.

* Assay: Name of the assay to be used for the pairwise analysis 
    (default='normalized')

* ... : Additional arguments to filter the analysis data by.

#### Boxplots and Lineplots steps

```{r BoxPlotsAndLinePlots}
################################################################################
### BoxPlots ###################################################################
################################################################################

subpathwayBoxplots(dat,
    subpathway = "Lactoyl Amino Acid", block_var = TIME1,
    treat_var = GROUP_NAME, Assay = "normalized",
    CHEMICAL_NAME = "CHEMICAL_NAME",
   SUB_PATHWAY = "SUB_PATHWAY", Gender == "Female"
)


################################################################################
## Line plots ##################################################################
################################################################################

# Set up data
dat$TIME1 <- as.numeric(factor(dat$TIME1,
    levels = c("PreSymp", "Onset", "End")
))

# Create line plots
subpathwayLineplots(dat,
    subpathway = "Lactoyl Amino Acid",
    block_var = TIME1, treat_var = GROUP_NAME,
    Assay = "normalized",
    CHEMICAL_NAME = "CHEMICAL_NAME",
   SUB_PATHWAY = "SUB_PATHWAY", Gender == "Female"
) +
    xlab("Time")
```

### Session Info
```{r Session Info }
sessionInfo()
```

