---
title: "MetabolomicsPipeline Workflow"
author: "Joel Parker"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 10,
  warning = F
)
```

```{r setup}
# Set global chunk obptions
knitr::opts_chunk$set(fig.width=12, fig.height=12, warning = F)

# Data 
library(readxl)
library(dplyr)

# Tables
library(table1)

# Plots
library(ggplot2)
library(ggplotify)

# Load Metabolomics Pipeline
library(MetabolomicsPipeline)

# To upload image
library(magick)

```

## Introduction

The purpose of the MetabolomicPipeline package is to provide additional tools to
complement the analysis done by Metabolon. This package allows us to create
a workflow that contains six different chunks.

1.  Peak normalization and standardization

2.  Analysis data creation

3.  Data exploration

4.  Subpathway Analysis

5.  Pairwise comparisons

6.  Box plot and line plots


## Data Description

In this vignette, we will use data which consists of 86 samples (42
males, 44 females), three treatment groups, and the samples were taken
at three different time points. In the original experiment, the sex of
each sample was not captured.
Therefore, we have an additional metadata Excel file which contains the
additional sex variable, as well as a variable called
"PARENT_SAMPLE_NAME, which allows us to link the extra metadata to
the sample metadata.

## Normalization and Standardization

Metabolon includes normalized peak data
as part of the Metabolon Excel file. We recommend you use the normalized
peak data from Metabolon; therefore, the normalization and
standardization steps are optional. However, in some cases, you may
need to perform normalization differently than Metabolon provided. An
example of this is experiments that include multiple batches.
Metabolon will perform the normalization for each batch separately, which
can cause downstream issues. When analyzing multiple batches we recommend combining the raw data and
performing one normalization on all batches.

In its raw form, the peak data contain counts for each metabolite in
each sample. Standardization and normalization are essential steps to
improve the signal-to-noise ratio. We implement normalization and standardization by:

1. Standardizing each metabolite by the metabolites median value.
    (median_standardization)

2. Imputing missing values for each metabolite by the minimum value of
    the metabolite. (min_val_impute)

3. Log transform each value (log_trans_met)

In the chunk below, we load the raw peak data from the Metabolon Excel
sheet and perform the normalization and standardization steps. Each
step only requires the output from the previous step.

```{r NormalizationAndStandardization, eval=FALSE}
###############################################################################
##### Load Data ###############################################################
###############################################################################
# Provide a path to Metabolon .xlsx file. 
metabolon_path <- "../data/UNAZ-0501-22VW_ DATA TABLES.xlsx" 

#  Load raw peak data
peak_data <- read_excel(metabolon_path, sheet = "Peak Area Data") 


###############################################################################
###### Standardization and Normalization ######################################
###############################################################################

# 1 Median standardization
med_std <- median_standardization(peak_data = peak_data)

# 2 Minimum Value imputation
impute <- min_val_impute(med_std)

# 3 Log Transformation
log_trans_met <- log_transformation(impute)
```

## Analysis Data Creation

We receive multiple datasets from Metabolon, each on a different tab of
the Metabolon Excel file. The datasets needed for the downstream
analysis are the sample metadata, chemical annotation, and
log-transformed peak data. The table below shows the tab that each
dataset is on within the Metabolon Excel file.

| Data                      | Tab of excel file    |
|---------------------------|----------------------|
| Sample metadata           | Sample Meta Data     |
| Chemical annotation       | Chemical Annotation  |
| Log-transformed peak data | Log Transformed Data |

The MetabolomicsPipeline package provides a convenient way to load each
of these datasets together using the "loadMetPipeFromMetabolon" function, which only
requires you to provide the path to the Metabolon Excel file.

In addition to the data from the Metabolon Excel file, we will also create an 
analysis dataset, which contains the analysis variables from the sample metadata
and the log-transformed peak data. 

#### Create Analysis Data Steps

We take the following steps to create the analysis data.

1. Load Data using the loadMetPipeFromMetabolon function.

 - We also load and add the additional metadata to the sample metadata.

2.  Enter metadata variables for analysis.

3.  Create the analysis data and add it to the rest of the experiment data. 

4.  Create a table showing the sample distribution.


```{r AnalysisDataCreation}
################################################################################
### Load Data ##################################################################
################################################################################

# Load Metabolon data from the Excel file
dat <- loadMetPipeFromMetabolon(metabolon_path = "../data/UNAZ-0501-22VW_ DATA TABLES.xlsx")

# load additional metadata
meta_data_additional <- read_excel("../data/AdditionalVars.xlsx")


# 2. Merge additional vars to the meta data
if(nrow(meta_data_additional)>0){
  
  meta_data <- meta_data_additional %>% 
    left_join(dat@meta,"PARENT_SAMPLE_NAME")
  
  # Update meta data slot
  dat@meta <- meta_data
}


################################################################################
### Enter Metadata Variables For Analysis ######################################
################################################################################
metadata_variables <- c("PARENT_SAMPLE_NAME", 
                        "GROUP_NAME",
                        "TIME1",
                        "Gender") 


################################################################################
### Add Analysis Data To MetPipe Object ########################################
################################################################################
# 2. Create analysis data
dat@analysis <- dat@meta %>% 
  select(all_of(metadata_variables)) %>% 
  left_join(dat@standardized_peak,"PARENT_SAMPLE_NAME") %>%
  as.data.frame()


################################################################################
### Create Table 1 #############################################################
################################################################################
# 5. Create table 1
tbl1 <- table1(~ GROUP_NAME + TIME1| Gender
               , data = dat@analysis) 

# 6. Display table 1
tbl1


# Save data
# saveRDS(dat, file = "data/dat.Rds")
```

## Exploratory Analysis

In data exploration, we use several methods to help us better understand
the underlying patterns in the data without using a formal hypothesis
test. In this pipeline, we are going to focus on two methods of data
exploration:

A.) Principal component analysis

B.) Heatmaps

#### Principal Component Analysis (PCA)

In general, Principal component analysis (PCA) reduces the number of
variables in a dataset while preserving as much information from the
data as possible. At a high level, PCA is constructed such that the
first principal component (PC) accounts for the largest amount of
variance within the data. The second PC accounts for
the largest remaining variance, and so on. Additionally, each
of the PCs produced by PCA is uncorrelated with 
the other principal components. PCA can allow us to visualize sources of
variation in the data. The metabolite_pca function will enable us to specify
a sample metadata variable to label the points in the plot.

Suppose you notice a variable with clearly separated groups that is not a variable of interest. In that case, consider
stratifying your downstream analysis by the values of that variable. For
example, we will stratify the downstream analysis by male/female in our vignette data. 

#### Heatmaps

For our heatmap, the
x-axis will be the samples, and the y-axis will be the metabolites. The
values determining the colors will be the log normalized peak values for each
metabolite in each observation. We can group the observations by the experimental
conditions. Grouping the experimental conditions in a heatmap is another way of visualizing sources of variation within our data.

We can use the metabolite_heatmap function to create the heatmaps, which requires the following arguments. 

* MetPipe: This is the experiment data

* top_mets: The number of metabolites to include in the heatmap. The metabolites are chosen based on the metabolites with the highest variability. 

* group_vars: The variables to group the samples by.
caption: The title of the heatmap.

* ... : You can add additional arguments to order the samples



#### Exploratory analysis steps

In the chunk below, we create a PCA plot labeled by Gender. Then, we make three heatmaps increasing by complexity. 

```{r ExploratoryAnalysis}
###############################################################################
### Run PCA ###################################################################
###############################################################################

# Define PCA label from metadata
meta_var = "Gender"

# Run PCA
pca <- metabolite_pca( dat,
               meta_var = meta_var)


# Show heatmap
pca

################################################################################
### Run Heatmaps ###############################################################
################################################################################

# Heatmap with one group
treat_heatmap <- metabolite_heatmap(dat,top_mets = 50,
                   group_vars = "GROUP_NAME",
                   strat_var = NULL,
                   caption = "Heatmap Arranged By Group",
                   GROUP_NAME)


as.ggplot(treat_heatmap)


# Heatmap with two groups
treatandtime <-  metabolite_heatmap(dat,top_mets = 50,
                   group_vars = c("GROUP_NAME","TIME1"),
                   strat_var = NULL,
                   caption = "Heatmap Arranged By Group and TIME",
                   GROUP_NAME, desc(TIME1))


as.ggplot(treatandtime)


# Heatmap with 2 group and stratified
 strat_heat <- metabolite_heatmap(dat,top_mets = 50,
                   group_vars = c("GROUP_NAME","TIME1"),
                   strat_var = "Gender",
                   caption = "Heatmap Arranged By Group and TIME",
                   GROUP_NAME, desc(TIME1))

 
## Female
as.ggplot(strat_heat[[1]])

# Male
as.ggplot(strat_heat[[2]])

```

## Subpathway Analysis

In the chemical annotation file, we will see that each metabolite is
within a sub-pathway, and each subpathway is within a superpathway.
There are several metabolites within each subpathway and several
sub-pathways within each Super-pathway. We can utilize an Analysis of
variance (ANOVA) model to test for a difference in peak intensities
between the treatment groups at the metabolite level, which is already
part of the Metabolon analysis. However, since multiple metabolites are
within a sub-pathway, it is challenging to test if the treatment
affected the peak data at the sub-pathway level. For this, we 
utilize a combined Fisher probability test. The combined Fisher test
combines the p-values from independent tests to test the hypothesis for
an overall effect. The Combined Fisher Probability is helpful
for testing a model at the subpathway level based on the pvalues from 
the model at the metabolite level.

### Combined Fished Analysis

We will test at the subpathway level by combining the p-values for each
metabolite within the subpathway for each model. We use a combination
function given by $\tilde{X}$ which combines the pvalues, resulting in a
chi-squared test statistic.

$$
\tilde{X} = -2\sum_{i=1}^k ln(p_i)
$$
where $k$ is the number of metabolites in the subpathway. We can
get a p-value from $P(X \geq\tilde{X})$, knowing that
$\tilde{X}\sim \chi^2_{2k}$. You will notice that smaller p-values will
lead to a larger $\tilde{X}$.

##### Assumptions

Since we are first testing each metabolite utilizing ANOVA, we make the
following assumptions for each metabolite,

-   *Independence:* Each observation is independent of all other
    observations. Therefore, if you have collected multiple samples from
    the same subject then this assumption may be violated.

-   *Normality:* The metabolite log-scaled intensities follow a normal
    distribution within each of the treatment groups.

-   *Homoscedasticity:* Equal variance between treatment groups.

In addition to the assumptions in the ANOVA models at the metabolite
level, the Fisher's Combined probability places an independence
assumption between the metabolites within the subpathway. 

For more about the Combined Fisher Probability and other methods that
can address this problem, see:

Loughin, Thomas M. "A systematic comparison of methods for combining
p-values from independent tests." Computational statistics & data
analysis 47.3 (2004): 467-485.

#### Models

To test our hypothesis at the subpathway level, we first have to form
our hypothesis at the metabolite level. For each metabolite, we test
three models.

1.) Interaction: $log Peak = Treatment Group + Time + Treatment*Time$

2.) Parallel: $log Peak = Treatment Group + Time$

3.) Single: $log Peak = Treatment$

For the interaction model, we are focusing only on the interaction term
"Treatment\*Time" to test if there is a significant
interaction between our treatment and the time variable. The parallel
model is testing if the time variable is explaining a significant
amount of the metabolite variance, and the treatment model is testing if
the treatment explains a significant proportion of the variance for each
metabolite.  

We test at the subpathway level using the Combined Fisher Probability method to combine the p-values from each model for all metabolites within the subpathway. To run the subpathway analysis, we use the "subpathway_analysis"
function, which requires the following arguments.  

* MetPipe: The experiment data.

* treat_var: The treatment variable of interest. 

* block_var: The block variable, in our example the block variable is going to be time.

* strat_var: The name of the variable we want to stratify our analysis by. In our 
this is going to be "Gender". 

#### Results Summaries

With the MetabolomicsPipeline package, we provide three different ways to summarize the results from the subpathway analysis.

1.  Number of significant subpathways by model type (subpath_by_model)

2.  Percentage of significant subpathways within superpathways (subpath_within_superpath)

3.  Metabolite model results within a specified subpathway (met_within_sub)


```{r SubpathwayAnalysis}
################################################################################
## Stratified Analysis #########################################################
################################################################################

# Stratified Analysis
stratified = subpathway_analysis(dat,
                                     treat_var = "GROUP_NAME",
                                     block_var = "TIME1",
                                     strat_var = "Gender")


################################################################################
### Results Plots ##############################################################
################################################################################

# 1. significant subpathways by model type
subpath_by_model(stratified)

# 2. Percentage of signficant subpathways within superpathways
subpath_within_superpath(stratified)

# 3. Metabolites within subpathway
tables <- met_within_sub(stratified, subpathway = "Partially Characterized Molecules")

### Females
tables[[1]]

### Males
tables[[2]]

```

## Pairwise Analysis

We can look at the pairwise comparisons for all
experimental groups at the metabolite level. Metabolon includes this as part of 
their analysis. However, if you need to change the model, you must rerun the pairwise
analysis. We will use the metabolite_pairwise function within the MetabolomicsPipeline
package, which requires the following arguments: 

* MetPipe: The experiment data.

* form: The pairwise_analysis function uses the model $log Peak = form$. Therefore, 
we must specify the right-hand side of the model. In our example, we will use
form = "GROUP_NAME + TIME1 + GROUP_NAME*TIME1".

#### Log Fold-Change Heatmap

We will produce a heatmap of the log fold changes for the metabolites with a significant overall p-value (which
tested if the treatment group means were equal under the null
hypothesis). The
heatmap colors will only show if the log fold-change is greater than
log(2) or less than log(.5). Therefore, this heatmap will only focus on
comparisons with a fold change of two or greater. The met_est_heatmap function will
produce an interactive heatmap using the results from the pairwise analysis.  

#### P-Value Heatmap

Similar to the pairwise estimate heatmap, we will produce a heatmap where the heatmap
will only include metabolites with a significant overall p-value, and the values
in the heat map will only be colored if the pairwise comparison is significant. We use the  
met_p_heatmap function to create an interactive p-value heatmap. 


```{r PairwiseAnalysis}
################################################################################
#### Run Pairwise Comparisons ##################################################
################################################################################

strat_pairwise = metabolite_pairwise(dat,form = "GROUP_NAME*TIME1",strat_var = "Gender")


###############################################################################
## Create Estimate Heatmap #####################################################
################################################################################

met_est_heatmap(strat_pairwise$Female, dat)


################################################################################
## Create P-value Heatmap ######################################################
################################################################################

# Female
met_p_heatmap(strat_pairwise$Female, dat)

```



## Boxplots and Lineplots

Visualizations of the data can help us see the underlying trends. Two
useful visualizations are boxplots and line plots, we will be using the subpathway_boxplots
and subpathway_lineplots functions to create them. The main utility of these functions
is it allows you for focus on the metabolites within a subpathway. For both functions, 
the arguments are:

* MetPipe: The analysis data

* subpathway: The name of the subpathway you would like for focus on

* X: The name of the variable for the X-axis

* groupBy: Name of the variable to group the results by. For us, we are going to
focus on the GROUP_NAME variable since we want to compare the treatment groups.  

* ... You can pass additional arguments to filter the results. In our example, we will focus on Gender=="Female" only.

#### Boxplots and Lineplots steps

```{r BoxPlotsAndLinePlots}
################################################################################
### BoxPlots ###################################################################
################################################################################

subpathway_boxplots(dat, subpathway = "Lactoyl Amino Acid", X=TIME1,
                    groupBy = GROUP_NAME, Gender =="Female")


################################################################################
## Line plots ##################################################################
################################################################################

# Set up data
dat@analysis$TIME1 <- as.numeric(factor(dat@analysis$TIME1,
                                         levels = c("PreSymp","Onset","End")))

# Create line plots 
subpathway_lineplots(dat, subpathway = "Lactoyl Amino Acid",
                     X= TIME1,groupBy = GROUP_NAME, Gender=="Female" )


```
